### 2024-11-14 paper/paper.qmd
Imported template

---
title: "Predicting Personal Total Income of Individuals in the United States in 2023"
subtitle: "XGBoost Captures the Complex Pattern Better Than Random Forest and Linear Regression"
author: 
  - Justin (Jiazhou) Bi and Weiyang Li
thanks: "All project related files available at: https://github.com/Jiazhou-Bi/ACS-Income-Model"
date: Nov 14, 2024
date-format: long
abstract: "This project aims to build a predictive model for personal total income using various demographic attributes, such as marital status, occupation, and residing state. We explored three popular machine learning algorithms: Random Forest, Linear Regression, and Extreme Gradient Boosting. Ultimately, Extreme Gradient Boosting was selected due to its superior predictive performance. While the model achieved a high R-squared value, it also exhibited a high Mean Squared Error, indicating challenges in accurately predicting all data points. Future steps for improving the model may involve incorporating additional demographic features to enhance model accuracy and reduce errors. Additionally, fine-tuning hyperparameters, experimenting with feature engineering, and handling outliers could further improve performance."

format: pdf
jupyter: python3 
number-sections: true
bibliography: references.bib
---

# Introduction {#sec-introduction}

Income is a fundamental concept as it affects nearly every member of society, influencing access to resources, quality of life, and overall economic stability. At the same time, it is also a difficult concept to define, as it can be interpreted differently depending on various perspectives, such as economic, social, legal, and tax purposes. Therefore, we must examine various demographical factors that may impact income to understand its concept better. Our project defines income as "all forms of net financial resources generated or lost from work, investments, or other activities." This project aims to predict personal income using data from the 2023 American Community Survey (ACS) [@ipums_usa_v15]. The ACS provides detailed demographic information, including age, education, employment, marital status, mortgage status, veteran status. As a result, it provides a robust dataset for studying potential factors that may influence income.

Due to the timeframe limitations, the findings of this project are specifically applicable to the 2023 dataset. As such, the results may vary across different periods, and their generalizability has not been assessed within this project. The raw data used in this analysis was sourced from IPUMS using the IPUMS API [@ipums_api]. We utilized pandas [@pandas] for data processing, and data visualization was carried out using Seaborn [@seaborn]. Additionally, all tables presented in this report were generated with Plotly [@plotly].

Income inequality is a serious problem and can lead to significant consequences and it may affect social structure, economic stability, as well as policy-making processes. Researchers found that demographic factors, such as education level and occupation type, can significantly impact an individual's financial standing [@Autor_2013]. Similarly, social relationships, such as marital status and household characteristics, affect income dynamics [@WaiteGallagher_2000]. Therefore, applying machine learning models to analyze the ACS dataset and using the information mentioned to predict personal income is reasonable.

We have chosen 2023 ACS data because it offers many demographical details, as well as self-reported total income of the individuals. The sample is also representative and covers all geographical locations in the USA. Furthermore, it provides easy access to their database through API. We use predictive models, including random forest, linear regression, and extreme gradient boosting machine to estimate personal income based on demographic factors to provide insights into which potential factors have the most significant impact on personal income. We hope the findings generated from our project can provide insights to policy-making process so that we can, hopefully, reduce income inequality and support further economic development in the future.

If we can successfully predict personal income using demographic information from the survey, this analysis will be able to contribute to a nuanced understanding of the American economy in 2023. It can help to reveal how demographic changess, educational attainment, and employment status/type can impact personal income. On the other hand, if the model cannot predict personal income with a high degree of confidence, this would suggest that the factors included in the analysiscannot not fully capture the intricacies of economic behavior and individual circumstances. As a result, additional variables, such as detailed education background like school of graduation, languages spoken, health status, etc, might be necessary in accurately predicting income.

In this analysis, we explored three different models—Linear Regression, Random Forest, and XGBoost—to predict income based on various demographic factors. Each model provides valuable insights into the relationships between income and predictors, such as education, age, industry, and gender. Among the models, XGBoost demonstrated the best performance in terms of predictive accuracy, as evidenced by its lower RMSE and superior handling of complex data interactions.

The remainder of this paper is structured as follows: @sec-data introduces the raw dataset and describes the cleaned datasets, along with a preliminary analysis through numerical summaries and visualizations. @sec-model describes the three machine learning models being used in our study. @sec-results explores key findings from our analysis. Lastly, @sec-discussion addresses the limitations of the analysis and offers recommendations for reducing delays and improving subway service reliability.

# Data {#sec-data}

The dataset used here is downloaded from the IPUMS USA website via [@ipums_usa_v15]. Our project's extracted dataset contains all the responses from the American Community Survey (ACS) from 2023. Our extraction included the following variables: total income, dwelling mortgage status, sex of the individual, marital status, education attainment, private or public school, industry or occupation, veteran status, and age. Some of the variables (such as educational level, occupation) we have chosen here have a more detailed version that contains more information. However, due to the nature of categorical variables, having detailed variables will create more classes of that variable, thus leading to more features for our model when encoding those variables. Although more features will improve the overall performance of our predictive model, it will also create issues such as overfitting, needing more computational resources, and lack of interpretability. We have chosen higher-level variables in this study for better computational efficiency. Furthermore, we have cleaned the data for better data quality. Details of our data cleaning processes are provided in @sec-datacleaning. In the following subsections, we will review all the variables used in this report and provide some basic descriptive statistics.

## Descriptive Data Analysis

In this subsection, we will explore every variable used in our study by examining some of the basic statistics. These preliminary analyses can help us better understand the data we are working with.

### Mortgage
The Mortgage variable describes the mortgage status of dwelling. There are five different values, they are "N/A", "No, owned free and clear", "Check mark on manuscript (probably yes)", "Yes, mortgaged/ deed of trust or similar debt", and "Yes, contract to purchase". We have dropped the N/A values for better data quality. Details available in @sec-datacleaning. The count of the remaining classes is shown below in (@fig-figure1).

```{python}

#| echo: false
#| label: fig-figure1
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Most Homeowners Are on Mortgage

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
mortgage_labels = {
    '1': 'No, owned free and clear',
    '3': 'Yes, mortgaged',
    '4': 'Yes, contract to purchase'
}
data['MORTGAGE'] = data['MORTGAGE'].astype(str).map(mortgage_labels)
plt.figure(figsize=(5, 3))
sns.countplot(x='MORTGAGE', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Mortgage Status', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='right')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

```

### Sex
This variable is coded as a binary variable, and we will use it as-is in our project. This does not suggest that the authors agree with this type of sex classification, but rather that it aligns with the available data and serves the research purposes of this project. The binary coding is a practical choice to maintain consistency with the dataset, though it may not fully capture the complexity and diversity of gender identities. @tbl-table1 shows the distribution of this variable.

```{python}

#| echo: false
#| label: tbl-table1
#| warning: false
#| message: false
#| fig-pos: "H"

import pandas as pd
import matplotlib.pyplot as plt

data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
sex_labels = {'1': 'Male', '2': 'Female'}
data['SEX'] = data['SEX'].map(sex_labels)
vet_status_count = data['SEX'].value_counts().reset_index()
vet_status_count.columns = ['Sex', 'Count']
fig, ax = plt.subplots(figsize=(6, 2))
ax.axis('off')
table = ax.table(
    cellText=vet_status_count.values,
    colLabels=vet_status_count.columns,
    cellLoc='center',
    loc='center'
)
table.auto_set_font_size(False)
table.set_fontsize(10)
table.scale(1, 1.5)
plt.tight_layout()
plt.show()

```

### Age
Age describes the age of the individual, and a age of 0 was coded for the individuals under 1 year old. For our dataset, the minimum age is 16, however. Age distribution of our dataset is shown below in @fig-figure2.
```{python}

#| echo: false
#| label: fig-figure2
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Age Distribution

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
plt.figure(figsize=(10, 6))
sns.countplot(x='AGE', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Age', fontsize=12)
plt.ylabel('Count', fontsize=8)
plt.xticks(rotation=90, ha='center')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

```

### Marital Status
Marital status describes the marital status of the individual. There are 7 different values, namely "Married, spouse present", "Married, spouse absent", "Separated", "Divorced", "Widowed", "Never married/single", and "Blank, missing". We have dropped the N/A values for better data quality. Details available in @sec-datacleaning. The following @fig-figure3 illustrate the distribution of marital status of the participants in our dataset.
```{python}

#| echo: false
#| label: fig-figure3
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Counts of Marital Status, Most Participants Are Married with Spouse Present, or Never Married/Single

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
marrital_status_label = {
  '1': 'Married, spouse present',
  '2': 'Married, spouse absent',
  '3': 'Separated',
  '4': 'Divorced',
  '5': 'Widowed',
  '6': 'Never married/single',
  '9': 'Blank, missing'
}
order = [
    'Married, spouse present',
    'Married, spouse absent',
    'Separated',
    'Divorced',
    'Widowed',
    'Never married/single'
]
data['MARST'] = data['MARST'].map(marrital_status_label)
plt.figure(figsize=(10, 6))
sns.countplot(x='MARST', data=data, order=order, palette='pastel', edgecolor='black')
plt.xlabel('Marital Status', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='center')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

```

### Educational Attainment
This is a general version (compared to the detailed version). The reason of choosing the general version is mentioned in @sec-data above. This column represents the highest educational attainment achieved by the individual. The distribution of the variable is shown in @fig-figure4.
```{python}

#| echo: false
#| label: fig-figure4
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Most Participants Have Middel School Attainment or Above

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
educ_new_label = {
    0: 'No Schooling',
    1: 'Nursery School',
    2: 'Primary School',
    3: 'Middle School',
    4: 'College 1-4 Years',
    5: 'College 5+ Years'
}
order = [
    'No Schooling',
    'Nursery School',
    'Primary School',
    'Middle School',
    'College 1-4 Years',
    'College 5+ Years'
]
data['EDUC_new'] = data['EDUC_new'].map(educ_new_label)
plt.figure(figsize=(10, 6))
sns.countplot(x='EDUC_new', data=data, order=order, palette='pastel', edgecolor='black')
plt.xlabel('Educational Attainment', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='center')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

```

### School Type
School type indicates if the individual went to a public school or private school. Other than missing values, there are three different classes. They are "Not enrolled", "Public school", and "Private school". We are not sure how the participants would respond if they have been to more than one type of schools. @fig-figure5 shows the distribution of our sample.
```{python}

#| echo: false
#| label: fig-figure5
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: School Type

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
school_type_label = {
    "1": 'Not Enrolled',
    "2": 'Private School',
    "3": 'Public School',
}
data['SCHLTYPE'] = data['SCHLTYPE'].map(school_type_label)
plt.figure(figsize=(10, 6))
sns.countplot(x='SCHLTYPE', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Private or Public School', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='center')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

```

### Industry
Industry describes the type of industry the individual works in. Same as the educational attainment variable, more detailed version of this variable is available (i.e., Occupation). There are more than 200 industries listed in our sample. A detailed description can be seen at [IPUMS USA](https://usa.ipums.org/usa-action/variables/IND1990#description_section).

### Veteran Status
Veteran status idicates whether the individual is a veteran or not. It can be either "Yes", "No", or "N/A". Please notice that "N/A" here is different from missing, suggesting this might be a different indicator, which is not explicitly explained from the ddl file provided. There is also a detailed version of this variable, which is not used in our project. The counts is displayed in @tbl-table2.
```{python}

#| echo: false
#| label: tbl-table2
#| warning: false
#| message: false
#| fig-pos: "H"

import pandas as pd
import matplotlib.pyplot as plt

data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
vet_status_labels = {'0': 'N/A', '1': 'No', '2': 'Yes'}
data['VETSTAT'] = data['VETSTAT'].map(vet_status_labels)
vet_status_count = data['VETSTAT'].value_counts().reset_index()
vet_status_count.columns = ['Veteran Status', 'Count']
fig, ax = plt.subplots(figsize=(6, 2))
ax.axis('off')
table = ax.table(
    cellText=vet_status_count.values,
    colLabels=vet_status_count.columns,
    cellLoc='center',
    loc='center'
)
table.auto_set_font_size(False)
table.set_fontsize(10)
table.scale(1, 1.5)
plt.tight_layout()
plt.show()
```

### Total Income
Total income is the total income made by the individual for the year. The distribution of total income is depicted in @fig-figure6. We have removed the outliers from the graph to have a better understanding of the general pattern of the overall distribution. From the figure, it is clear that more income are reported near multiple of $10,000 as a unit. This may be caused by the "heaping and leaping" phenomonon found in self-reported survey data [@pudney2008heaping].
```{python}

#| echo: false
#| label: fig-figure6
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Total Income Distribution

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
plt.figure(figsize=(10, 6))
sns.histplot(x='INCTOT', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Total Income', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='center')
plt.xlim(0, 250000)
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

```

## Analysis of Feature Relationships {#sec-afr}

In this subsection, we will explore some of the potential relationships among the features chosen for our study. Understanding these relationships can prepare us better for our model construction in the next section.

### Income Distribution by Mortgage Status

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../other/graphs/EDAincomebymortgage.png}
    \caption{Income Distribution by Mortgage Status}
    \label{fig:income-mortgage}
\end{figure}

Figure \ref{fig:income-mortgage} illustrates the income distribution for individuals with different mortgage statuses. The highest average income is observed among individuals with a "Mortgage/Trust Deed," followed by those who own their homes outright ("Fully Owned"). Those with a "Signed Purchase Contract" have the lowest average income among the groups. This pattern may reflect a correlation between mortgage status and income level, where individuals with higher incomes are more likely to secure traditional mortgage loans, while those with lower incomes tend towards alternative purchasing arrangements. This could indicate the influence of financial stability and access to credit on mortgage status.

### Correlation Analysis of Age, Education, and Income

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../other/graphs/EDAcorrelation.png}
    \caption{Correlation Matrix of Age, Education, and Income}
    \label{fig:correlation-matrix}
\end{figure}

Figure \ref{fig:correlation-matrix} shows the correlation matrix for three key variables: age, education level (EDUC_new), and total personal income (INCTOT). The matrix reveals that there is a moderate positive correlation between education level and income (0.29), suggesting that higher education levels are generally associated with higher income. Additionally, a smaller positive correlation exists between age and income (0.14), which might indicate that as individuals grow older, they tend to earn more due to accumulated experience. The correlation between age and education (0.06) is minimal, which is expected as these two variables are less directly related. Overall, this analysis supports the notion that education level plays a more significant role in income variation than age within this dataset.

### Average Income by Education and Gender {#sec-afr3}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../other/graphs/EDAinteraction.png}
    \caption{Average Income by Education and Sex}
    \label{fig:income-education-sex}
\end{figure}

Figure \ref{fig:income-education-sex} presents the average total personal income based on different levels of education for both male and female groups. From the plot, it is observed that individuals with higher levels of education, particularly those who attended "College 5+ Years," tend to earn more on average compared to other groups. However, there are some variations in income within each education level, as indicated by the error bars. Generally, males tend to have higher average incomes across most education levels. This difference suggests a possible gender-based income disparity, even after controlling for education levels. Additionally, the variability in income among education levels highlights that factors beyond educational attainment, such as occupation and industry, may further impact income. Thus, incorporating an interaction term between education and gender in the model is necessary to capture these nuanced effects more accurately.

# Model {#sec-model}

Our modeling strategy aims to explore the relationship between personal characteristics and income. We implemented three models: Linear Regression with Interaction Terms, Random Forest, and XGBoost, each chosen for their unique strengths in prediction and interpretation.

## Train Test Split of Data

We have cleaned our data before we split it into training and testing datasets. We are aware of potential data leakage of our testing data, and we have mitigated this risk by not using overall statistics or normalization that requires the original data to calculate, not imputing any missing data, or engineering features that include the original dataset. By doing so, we can minimize the risk of data leakage, and ensure the perfect seventy and thirty percent train test split.

## Linear Regression Model with Interaction Term {#sec-linear-regression-model-with-interaction-term}

### Model set-up

We implemented a linear regression model with an interaction term to explore the relationship between personal characteristics and income. The general form of the model is:

$$
y=\beta_0+\beta_1X_1+\beta_2X_2+...+\beta_nX_n+\beta_{\text{interaction}}(X_i\times X_j)+\epsilon
$$

where:

-   y represents the total income (`INCTOT`),

-   $X_1, X_2, ... , X_n$ are the predictor variables, including both categorical and numerical features,

-   $X_i \times X_j$ is the interaction term between education and sex (`EDUC_SEX_INTERACTION`)

-   $\beta_0$ is the intercept,

-   $\beta_1, \beta_2, ... , \beta_n, \beta_{\text{interaction}}$ are the coefficients for the respective predictors and interaction term.

To ensure the robustness of our linear regression model, we conducted a series of steps aimed at minimizing multicollinearity and improving model interpretability. Initially, we included all relevant predictor variables, both categorical and numerical, to explore the full range of relationships between personal characteristics and income. Next, we calculated the Variance Inflation Factor (VIF) for each predictor to identify potential multicollinearity issues. Typically, a VIF value greater than 10 indicates problematic multicollinearity, which can distort the model's estimates. In our analysis, variables such as `VESTAT` (Veteran Status) and `AGE` had VIF values exceeding 13, suggesting they were highly correlated with other predictors. To address this, we removed these high-VIF variables from the model. Afterward, we refit the linear regression model using the reduced set of predictors, which resulted in improved performance metrics, including a lower Mean Squared Error and a more interpretable set of coefficients.

Note that according to our previous analysis in @sec-afr3, the interaction term, `EDUC_SEX_INTERACTION`, was included to capture the combined effect of education level and gender on income. This allowed us to model the non-linear relationship between these two variables and their joint impact on the outcome variable.

::: {#tbl-coefficients}
\begin{center}
Table 2: Coefficients from the Linear Regression Model
\end{center}

| Feature                     | Coefficient |
|-----------------------------|-------------|
| cat__MORTGAGE             | 11267.90    |
| num__AGE                  | 415.91      |
| cat__STATEICP             | 126.39      |
| cat__IND1990              | -43.18      |
| cat__OCC2010              | -139.55     |
| num__EDUC_SEX_INTERACTION | -1832.32    |
| cat__MARST                | -5541.39    |
| cat__GQ                   | -7956.89    |
| cat__SCHLTYPE             | -16320.74   |
:::

*Table 2: Linear Regression Coefficients Table.*

We run the model in Python [@cite_python] using the `scikit-learn` package [@cite_sklearn]. We use the default settings for the `LinearRegression` function from `scikit-learn`. We use the `statsmodels` package [@cite_statsmodels] to calculate the VIF for each predictor.

### Model Justification

In evaluating our model's performance, we selected key metrics to assess both its fit and predictive accuracy, including the Variance Inflation Factor (VIF), R-squared ($R^2$), and Mean Squared Error (MSE). These metrics were chosen to provide a comprehensive evaluation of the model's robustness and reliability in capturing the relationship between personal characteristics and income.

The VIF was used to diagnose multicollinearity among predictors, helping us ensure that no variables were highly correlated, which could distort the interpretation of the model's coefficients.

::: {#tbl-VIF}
\begin{center}
Table 3: Variance Inflation Factors (VIF) for Predictor Variables
\end{center}

| Feature                  | VIF       |
|--------------------------|-----------|
| cat__STATEICP            | 3.24  |
| cat__GQ                  | 1.00  |
| cat__MORTGAGE            | 2.66  |
| cat__MARST               | 1.74  |
| cat__SCHLTYPE            | 1.24  |
| cat__OCC2010             | 3.19  |
| cat__IND1990             | 8.49  |
| num__AGE                 | 7.87  |
| num__EDUC_SEX_INTERACTION| 6.95  |
:::

*Table 3: This table displays the Variance Inflation Factors (VIF) for each of the predictor variables in the model. VIF values greater than 10 indicate potential multicollinearity issues.*

We chose $R^2$ as it provides a measure of how well the model explains the variance in the dependent variable, making it an important indicator of the model’s explanatory power. The MSE was selected to quantify the average squared difference between observed and predicted values, offering a clear view of the model’s prediction error.

::: {#tbl-modelperformance}
\begin{center}
Table 4: Model Performance Metrics
\end{center}

| Metric                     | Value              |
|----------------------------|--------------------|
| R-squared                  | 0.0774       |
| Adjusted R-squared         | 0.0774     |
| Mean Squared Error (MSE)   | 7,448,800,507.01   |
| Root Mean Squared Error (RMSE) | 86,306.43        |
:::

*Table 4: This table presents the performance metrics of the linear regression model, including MSE, RMSE, R-squared, and Adjusted R-squared. These metrics are used to assess the model's fit and predictive accuracy.*

Additionally, residual analysis was conducted by plotting residuals against fitted values to check for randomness around zero, ensuring that the model satisfied the assumptions of linearity and homoscedasticity.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.5\textwidth]{../other/graphs/LinearModelResidual1.png}
\caption{Linear Model Residuals Plot}
\label{fig:residual-linear1}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width=0.5\textwidth]{../other/graphs/LinearModelResidual2.png}
\caption{Linear Model Residuals vs. Predicted Values Plot}
\label{fig:residual-linear2}
\end{figure}

We use the statsmodels package [@cite_statsmodels] to calculate the VIF for each predictor. We use the matplotlib [@cite_matplotlib] and seaborn [@seaborn] to visualize the residuals. We use the scikit-learn [@cite_sklearn] package to calculate the $R^2$ score.

## Random Forest Model {#sec-random-forest}

### Model set-up

To analyze the non-linear relationship between personal characteristics and income, we employed a Random Forest model. Random Forest is an ensemble learning method that builds multiple decision trees and averages their predictions, reducing overfitting and capturing complex, non-linear relationships in the data. We utilized Bagging (Bootstrap Aggregating) and random feature selection to enhance model performance and reduce overfitting. Bagging involves generating multiple bootstrap samples from the original training set, with each sample used to train an independent decision tree. By averaging the predictions of multiple trees, Bagging reduces the variance of the model, thereby improving its robustness. Additionally, Random Forest introduces further randomness by selecting a subset of features at each split, which reduces correlation among trees and improves the model’s generalization ability.

To optimize the model, we conducted hyperparameter tuning, focusing on two key parameters: `n_estimators` (the number of trees) and `max_depth` (the maximum depth of each tree). Increasing `n_estimators` typically decreases model variance and improves stability, but too many trees can result in diminishing returns with increased computation time. Initially, we set `n_estimators` to 100, but through Grid Search, we tested values of 100, 200, and 300 and found that 300 trees yielded the best performance. Similarly, the `max_depth` parameter controls the complexity of the trees. While deeper trees can capture more intricate data patterns, they may also overfit. We initially set `max_depth` to 15 and test 10, 15, 20 but found that reducing it to 10 during hyperparameter tuning improved generalization by avoiding overfitting. Additionally, we tuned `min_samples_split` and `min_samples_leaf` to prevent overfitting by ensuring that nodes have a sufficient number of samples before splitting.

We run the model in Python [@cite_python] using the scikit-learn package [@cite_sklearn] and pandas [@pandas]. We used the GridSearchCV function from scikit-learn for hyperparameter tuning and subsampled the training data using pandas. The RandomForestRegressor function from scikit-learn was employed for model fitting.

### Model Justification

To assess the performance of our Random Forest model, we evaluated several key metrics, including Mean Squared Error (MSE), Root Mean Squared Error (RMSE), feature importance, and residual analysis. These evaluations provided insights into the model’s predictive accuracy, interpretability, and overall fit.

First, we calculated the Mean Squared Error (MSE) and Root Mean Squared Error (RMSE) to assess the overall prediction accuracy of the model. These metrics provided insight into how well the model captured the variance in income based on the selected features.

::: {#tbl-randomforestMSE}
\begin{center}
Table 5: Best Random Forest Model Performance Metrics
\end{center}

| Metric          | Value            |
|-----------------|------------------|
| Best Model MSE  | 6,057,426,411.65 |
| Best Model RMSE | 77,829.47        |
:::

*Table 5: This table shows the Mean Squared Error (MSE) and Root Mean Squared Error (RMSE) for the best-tuned Random Forest model, demonstrating the model's performance after hyperparameter tuning.*

Additionally, we performed a feature importance analysis, which allowed us to understand the relative significance of each predictor variable in determining income. This is particularly useful for verifying that the model identifies the key factors influencing income as expected.

::: {#tbl-randomforestimportance}
\begin{center}
Table 6: Feature Importance in the Best Random Forest Model
\end{center}

| Feature         | Importance |
|-----------------|------------|
| cat\_\_EDUC_new | 0.337   |
| cat\_\_IND1990  | 0.243   |
| num\_\_AGE      | 0.194   |
| cat\_\_SEX      | 0.149   |
| cat\_\_MARST    | 0.056   |
| cat\_\_MORTGAGE | 0.013   |
| cat\_\_SCHLTYPE | 0.004   |
| cat\_\_VETSTAT  | 0.003   |
:::

*Table 6: This table lists the feature importance values for the best-performing Random Forest model. The higher the importance value, the more significant the feature is in predicting the target variable (income).*

We also conducted a residual analysis by plotting the residuals (the difference between actual and predicted values) against the predicted values. This step helped us evaluate whether the model appropriately captured the underlying patterns in the data, ensuring that there were no significant biases or violations of model assumptions.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.5\textwidth]{../other/graphs/RandomForestResidual.png}
\caption{Random Forest Model Residuals vs. Predicted Values Plot}
\label{fig:residual-randomforest}
\end{figure}

The feature importances were extracted from the best model using RandomForestRegressor from scikit-learn [@cite_sklearn] and displayed using pandas [@pandas]. We use the matplotlib [@cite_matplotlib] to visualize the residuals.

## XGBoost {#sec-xgboost}

### Model set-up

XGBoost (Extreme Gradient Boosting) is a powerful machine learning algorithm that builds an ensemble of decision trees using gradient boosting techniques. In gradient boosting, each subsequent tree is trained to correct the residuals of the previous trees, iteratively improving the model’s predictive accuracy. XGBoost enhances this process by incorporating regularization techniques to control overfitting, making it an ideal choice for capturing complex, non-linear relationships in data.

To optimize the performance of our XGBoost model, we employed GridSearchCV to conduct hyperparameter tuning. We explored a range of hyperparameters to find the optimal configuration for our dataset. The grid search included the learning rate (`learning_rata`), max depth (`max_depth`), number of trees (`n_estimators`), and minimum child weight (`min_child_weight`). Specifically, we tested learning rates of 0.01, 0.05, and 0.1, tree depths of 4, 6, and 8, and `n_estimators` values of 100, 200, and 300. Using GridSearchCV with a cross-validation fold of 3, we systematically evaluated different combinations of these hyperparameters. The model was evaluated using negative Mean Squared Error (MSE) as the scoring metric. The optimal model used a learning rate of `0.1`, a maximum tree depth of `6`, a minimum child weight of `5`, and `300` boosting rounds. This configuration provided a good balance between model complexity and predictive accuracy, capturing important non-linear relationships while minimizing overfitting.

We implemented the XGBoost model in Python [@cite_python] using the scikit-learn package [@cite_sklearn], with pandas [@pandas] for data processing and xgboost [@cite_xgboost] for model fitting. The initial model was trained using default parameters of the XGBRegressor function, which was subsequently fine-tuned using GridSearchCV from scikit-learn [@cite_sklearn] for hyperparameter optimization.

### Model Justification

Similar to the Random Forest model, the XGBoost model was trained on a 70-30 train-test split. We evaluated the model's performance using Mean Squared Error (MSE) and Root Mean Squared Error (RMSE) to assess its predictive accuracy.

::: {#tbl-XGBoostMSE}
\begin{center}
Table 7: XGBoost Model Performance Metrics
\end{center}

| Metric                         | Value            |
|--------------------------------|------------------|
| Mean Squared Error (MSE)       | 5,854,629,848.13 |
| Root Mean Squared Error (RMSE) | 76,515.55        |
:::

*Table 7: This table shows the Mean Squared Error (MSE) and Root Mean Squared Error (RMSE) for the final XGBoost model, reflecting its predictive accuracy.*

We performed a residual analysis by plotting the residuals against the predicted values to assess the model's fit. This allowed us to check whether the XGBoost model accurately captured the patterns in the data and to verify that no major biases or violations of model assumptions were present.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.5\textwidth]{../other/graphs/XGBoostResidual.png}
\caption{XGBoost Model Residuals vs. Predicted Values Plot}
\label{fig:residual-XGBoost}
\end{figure}

We also performed cross-validation to ensure the robustness of the XGBoost model. This technique allowed us to evaluate the model’s performance across different subsets of the data, helping to confirm that the model generalizes well and is not overly sensitive to any particular portion of the training data.

::: {#tbl-XGBoostCV}
\begin{center}
Table 8: XGBoost Cross-Validation Performance
\end{center}

| Fold  | RMSE       |
|-------|------------|
| 1     | 76,693.86  |
| 2     | 75,692.41  |
| 3     | 76,193.42  |
| 4     | 77,365.82  |
| 5     | 76,552.10  |
| **Average** | **76,499.52** |
:::

*Table 8: This table shows the Root Mean Squared Error (RMSE) from cross-validation, demonstrating the model’s performance across different subsets of the data.*

We used matplotlib [@cite_matplotlib] to perform a residual analysis against the predicted values from the XGBoost model. We performed cross-validation using cross_val_score function from scikit-learn  [@cite_sklearn] to evaluate the robustness of the model.

# Results {#sec-results}

## Linear Regression Model with Interaction Term

### Linear Regression Results

Table \ref{tbl-coefficients} provides the coefficients for each predictor in the model. The coefficients represent the estimated change in the target variable (income) for a one-unit change in the predictor, holding all other variables constant.

The largest positive effect on income is observed for `cat_MORTGAGE`, which has a coefficient of 11,267.90. This suggests that individuals with a mortgage tend to have significantly higher incomes compared to those without, when other factors are held constant. Additionally, the coefficient for `num__AGE` is 415.91, indicating that income tends to increase with age, though the effect size is relatively small.

On the other hand, some predictors have negative effects on income. The variable `cat__SCHLTYPE` (school type) has the largest negative coefficient of -16,320.74, suggesting that attending certain types of schools might be associated with lower income levels. Similarly, `cat__GQ` and `cat__MARST` (marital status) have negative coefficients of -7,956.89 and -5,541.39, respectively, implying that certain marital statuses or living arrangements are associated with lower income levels.

Interestingly, the interaction term `num__EDUC_SEX_INTERACTION` has a negative coefficient of -1,832.32, suggesting that the relationship between education and income may differ by gender. The negative value could indicate that the combined effect of education and gender has a suppressing effect on income for certain groups.

### Model Performance

As shown in Table \ref{tbl-modelperformance}, the linear regression model explains a small portion of the variance in income, as evidenced by the R-squared value of 0.0775, which indicates that only 7.75% of the variability in income is explained by the predictors in the model. The Adjusted R-squared, which adjusts for the number of predictors, is slightly lower at 0.0775, confirming that the inclusion of additional predictors did not drastically improve the model’s explanatory power.

Furthermore, the model’s Mean Squared Error (MSE) is 7,448,800,507, and the Root Mean Squared Error (RMSE) is 86,306.43, suggesting that the model's predictions deviate significantly from the actual income values. These performance metrics indicate that the linear regression model has limited predictive accuracy for this dataset.

### Residual Analysis

The residual analysis helps evaluate the linear regression model's performance. The residual histogram as shown in Figure \ref{fig:residual-linear1} shows a strong right skew, indicating that while most predictions are close to the actual values, the model struggles with higher incomes. The long tail suggests significant under-prediction for some individuals.

In the residuals vs. predicted values plot as shown in Figure \ref{fig:residual-linear2}, the residuals exhibit a funnel shape, showing increasing variance as predicted income rises. This indicates heteroscedasticity, meaning the model's error increases for higher predicted values. Overall, the analysis suggests that the linear model fits well for lower incomes but underperforms for higher-income predictions, implying a need for more complex models to better capture these relationships.

## Random Forest

We employed a Random Forest model to predict income, tuning several hyperparameters to improve model performance. The best model was achieved with the following parameters: **max_depth=10**, **min_samples_leaf=4**, **min_samples_split=10**, and **n_estimators=300**. As we shown in Table \ref{tbl-randomforestMSE}, the final model had a Mean Squared Error (MSE) of 6,057,426,411.65 and a Root Mean Squared Error (RMSE) of 77,829.47, indicating a substantial improvement in prediction accuracy compared to the linear regression model.

### Feature Importance

The Random Forest model, Table \ref{tbl-randomforestimportance}, also allows for an examination of feature importance, which ranks the variables by their contribution to the model's predictions. The most important features in the model are:

-   `cat__EDUC_new` (Education level) with an importance score of 0.337, indicating that education is the most significant factor affecting income in this model.

-   `cat__IND1990` (Industry) has an importance score of 0.243, suggesting that the industry in which a person works is also a major determinant of income.

-   `num__AGE` (Age) comes in third with a score of 0.194, implying that age plays a substantial role in income prediction.

Other variables, such as `cat__SEX` (Sex) and `cat__MARST` (Marital Status), show moderate importance, while factors like `cat__MORTGAGE` and `cat__SCHLTYPE` (School Type) have much smaller impacts.

### Residual Analysis

The residuals from the Random Forest model as shown in Figure \ref{fig:residual-randomforest} were analyzed similarly to the linear model. The residuals plot, comparing predicted values with residuals, shows less heteroscedasticity compared to the linear model, suggesting that the Random Forest model handles the data more effectively, particularly for higher income levels. However, some degree of variability in the residuals still increases with predicted income, indicating that further tuning or model adjustments may be necessary for optimal performance.

## XGBoost

The XGBoost model, a gradient boosting algorithm, was optimized using a hyperparameter grid search, yielding the best parameters: **learning_rate=0.1**, **max_depth=6**, **min_child_weight=5**, and **n_estimators=300**. These parameters were selected to balance model complexity and overfitting control. As we shown in Table \ref{tbl-XGBoostMSE}, the final model achieved a Mean Squared Error (MSE) of 5,854,629,848.13 and a Root Mean Squared Error (RMSE) of 76,515.55, making it the best-performing model among the three evaluated.

### Model Performance

The XGBoost model’s Cross-Validation RMSE (Table \ref{tbl-XGBoostCV}) was 76,499.52, which aligns closely with the final RMSE, indicating that the model generalizes well across different data subsets. This consistency in performance suggests that XGBoost effectively captures the relationship between predictors and income without overfitting.

### Residual Analysis

The residuals vs. predicted values plot as shown in Figure \ref{fig:residual-XGBoost} shows a similar pattern to the Random Forest model, with residuals clustering around zero for lower predicted values but spreading out for higher predicted incomes. However, compared to the Random Forest model, the XGBoost residuals show a slightly narrower spread, particularly at the higher income levels, indicating that XGBoost handles high-income predictions somewhat better. Nonetheless, some degree of heteroscedasticity remains, as the spread increases with predicted income.

# Discussion {#sec-discussion}

In this analysis, we explored three different models—Linear Regression, Random Forest, and XGBoost—to predict income based on various demographic and categorical factors. Each model provided valuable insights into the relationship between income and predictors such as education, age, industry, and gender. Among the models, XGBoost demonstrated the best performance in terms of predictive accuracy, as evidenced by its lower RMSE and superior handling of complex data interactions. However, this does not mean it is without limitations or that further improvements cannot be made.

## Key Findings

### Feature Importance
We have found that educational attainment and occupation industry are consistently the most important features in all three models, especially in XGBoost and Random Forest. This indicates that higher educational attainment and certain industries are correlated with higher personal income. This confirms our original hypothesis that certain demographic factors, such as education, may lead to specialized industry sectors where more income can be generated.

Age is also a useful predictor. However, despite the common belief that age is a determining factor when predicting personal income, its effectiveness is not as strong as education or industry. This may be caused by the fact that in some industries, your experience may hit a plateau, after which age may not contribute as much to experience as before certain age.

Marital status and sex do affect personal income, but not as significantly as the other features. It is great to prove that certain demographical features are not as informative when predicting personal income, suggesting a "fair" distribution of wealth.

### Models

None of the three models used in our study provided an excellent prediction of personal income. For the residual analysis, it appears that the models' predictbility deterioate as the self-reported income gets higher. REsidual variance increased as the predicted income increased in the Random Forest and XGBoost model, suggesting potential heteroscedasticity. Our educated guess is that extra-high income may come from various sources, and simplifying them into a simple model like this is not a reasonable solution.

Among the three models, XGBoost has the highest accuracy overall. We believe that it can capture more complex relationships and interactions among the variables than the other two models. However, the high MSE for extra-high income individuals indicates that the model still needs to be refined more to achieve an accurate prediction.

Linear Regression has underperformed relative to the other two models. We speculate that simplicity in modeling linear relationships did not provide sufficient information to capture the non-linear patterns underlying our data, especially where multiple features interact to influence income, as suggested in @sec-afr.

## Weaknesses and Limitations

One major limitation encountered throughout the analysis was the skewness of the income distribution. The income data is heavily left-skewed, with most individuals earning relatively low amounts and a few individuals earning substantially higher incomes. This skewness likely contributed to the residual patterns we observed, particularly in the linear and Random Forest models.

We attempted to address this by applying a logarithmic transformation to income in an effort to normalize the data. However, this transformation did not lead to substantial improvements in model performance. Even after the transformation, the residual patterns remained, indicating that the underlying relationships between predictors and income might require more sophisticated modeling approaches than simple transformations can provide.

Another potential solution—though not ideal—would be to exclude extreme income values from the analysis. Removing high-income outliers might reduce the heteroscedasticity issue and result in a better fit for the majority of the population. However, excluding high-income individuals would also mean losing valuable information about the upper end of the income distribution, which is crucial for a complete analysis. Before taking this step, it would be necessary to justify it based on domain-specific knowledge or further data exploration.

## Next Steps

Looking forward, there are several potential avenues for further research and model improvement:

-   While we focused on standard predictors like age, education, and industry, additional variables or interaction terms might improve model performance. For example, interactions between age and education, or industry and gender, could capture more nuanced relationships.

-    Instead of removing high-income individuals, more sophisticated outlier treatment techniques could be applied. For instance, applying quantile regression or using robust regression models might help better model extreme values without distorting the broader data patterns.

-   The skewness in income distribution poses a significant challenge for modeling. Exploring techniques like synthetic data generation for high-income individuals or applying weighting schemes to account for data imbalance might help the models better capture the variance across all income levels.

In conclusion, while XGBoost provided the most accurate results, there is room for improvement in handling skewed data and better modeling high-income individuals. Future work will focus on addressing these issues while exploring more advanced modeling techniques and feature engineering to enhance predictive performance further.

\newpage

# Appendix

## Data Cleaning {#sec-datacleaning}

In the data cleaning process, we first examined all required variables and removed any observations with missing values (N/A). Notably, the variable `CITY`, which we initially intended to include in the analysis, was excluded due to its high proportion of missing values, which compromised data completeness.

For the response variable, personal total income (INCTOT), we identified several non-informative coded values, such as `0000000 = None`, `9999998 = Unknown`, and `-009995 = -$9,900 (1980)` and so on. These values lack meaningful information for analysis, so we removed any records containing such entries to maintain the integrity of the income data.

In terms of predictor variables, we found that several included codes like "not applicable" or "does not exist," which represent cases without relevant data for our analysis. These values were also removed to ensure only relevant data points were used in modeling.

Furthermore, we simplified the `EDUC` variable (education level), which originally included highly granular categories (e.g., fifth grade, sixth grade, second year of college, fifth year of college). Such detailed classifications were unnecessary for our purposes, so we re-coded `EDUC` into a new variable, `EDUC_new`, with broader categories:

-   'No Schooling': 0,
-   'Nursery School': 1,
-   'Primary School': 2,
-   'Middle School': 3,
-   'College 1-4 Years': 4,
-   'College 5+ Years': 5

This simplified categorization enhances interpretability and usability in the analysis. By conducting these data cleaning steps, we produced a refined and consistent dataset, ready for accurate and reliable modeling.

\newpage

# References

### 2024-11-20 paper/paper.qmd
Abstract updated. Sketches updated.

@@
 ---
-title: "Predicting Personal Total Income of Individuals in the United States in 2023"
-subtitle: "XGBoost Captures the Complex Pattern Better Than Random Forest and Linear Regression"
+title: "Comparing Different Machine Learning Models at Predicting Bitcoin Price Moving Direction at Different Time Intervals from 2017 to 2024"
+subtitle: "RNN Can Predict Best at One-Hour Intervals"
 author: 
-  - Justin (Jiazhou) Bi and Weiyang Li
-thanks: "All project related files available at: https://github.com/Jiazhou-Bi/ACS-Income-Model"
-date: Nov 14, 2024
+  - Justin (Jiazhou) Bi
+thanks: "All project related files available at: https://github.com/Jiazhou-Bi/BTC-Price-Prediction"
+date: Nov 20, 2024
 date-format: long
-abstract: "This project aims to build a predictive model for personal total income using various demographic attributes, such as marital status, occupation, and residing state.
-We explored three popular machine learning algorithms: Random Forest, Linear Regression, and Extreme Gradient Boosting.
-Ultimately, Extreme Gradient Boosting was selected due to its superior predictive performance.
-While the model achieved a high R-squared value, it also exhibited a high Mean Squared Error, indicating challenges in accurately predicting all data points.
-Future steps for improving the model may involve incorporating additional demographic features to enhance model accuracy and reduce errors.
-Additionally, fine-tuning hyperparameters, experimenting with feature engineering, and handling outliers could further improve performance."
+abstract: "This project aims to build a predictive model for the price moving direction of Bitcoin (BTC) at 1-minute, 1-hour, and 1-day intervals.
+We explored four popular machine learning algorithms for time-series data prediction: Recurrent Neural Network (RNN), Long Short-Term Memory (LSTM), Random Forest (RF), and Extreme Gradient Boosting (xGB).
+We have compared these four models on two different dimensions: model performance and time for training.
+The data used for this project is gathered from Binance which contains BTC data from late 2017 to present (November 2024).The original variables include high, low, open, close, and volume, as well as the timestamp of BTC, at one-minute, one-hour, and one-day intervals.
+Ultimately, RNN has the overall best performance in terms of its predictive accuracy, while XGB has the fastest training duration.
+While most models across different time intervals achieved a better performance than simply guessing at a 50-50 percent chance, their performance varies for each direction of the movemenet.
+Future steps for improving the model may involve incorporating additional features to enhance model accuracy and reduce errors.
+Additionally, fine-tuning hyperparameters, experimenting with more sophisticated feature engineering, and handling outliers could further improve performance."
 
@@
 # Introduction {#sec-introduction}
+
+Cryptocurrencies have drawn much attention of the public recently due to its growing public awareness.
+xxx amount of papers/news about BTC was published (refereces here) from xxxx to xxxx.
+According to xxx (reference here), the value of the market...
+etc.
+The trading volume (references) ...
+It has been adopted by xxx platforms and ....
+(reference) Also, mention the importance of BTC (references needed) and its role of being a indicator of the overall crypto market.
+Therefore, it has become one of the most traded ....
+On the other hand, due to its non-tangible nature, volitility is key aspect.
+xxxx data about its ups and downs.
+Only contains data from 2017 because ....
+fits our scope.
+Regain values (references 5 in the paper).
+Similar to the stock prices, because BTC is considered a tradable investment instrument, many attempt to predict its price in the future to faciliate their trading strategy.
+Recently studies were published to discuss whether the price of BTC is predictable and technical indicators returned the possibility.
+Accuracy up to 63% for direction prediction (reference here, 9).
+Also, in 10, they reported 98% success rate for daily price forecast.
+Based on recent data 2017 and 2018.
+Both EOD prediction.
+We have even short trading horizons, 1 min, 1 hour.
 

### 2024-12-05 paper/old paper.qmd
EDA graphs saved. Abstract, Intro, Data completed. LFS tracks all parquet files now.

--- 2024-11-20 paper/paper.qmd
@@
 thanks: "All project related files available at: https://github.com/Jiazhou-Bi/BTC-Price-Prediction"
-date: Nov 20, 2024
+date: Dec 3, 2024
 date-format: long
@@
 Cryptocurrencies have drawn much attention of the public recently due to its growing public awareness.
-xxx amount of papers/news about BTC was published (refereces here) from xxxx to xxxx.
-According to xxx (reference here), the value of the market...
-etc.
-The trading volume (references) ...
-It has been adopted by xxx platforms and ....
-(reference) Also, mention the importance of BTC (references needed) and its role of being a indicator of the overall crypto market.
-Therefore, it has become one of the most traded ....
-On the other hand, due to its non-tangible nature, volitility is key aspect.
-xxxx data about its ups and downs.
-Only contains data from 2017 because ....
-fits our scope.
-Regain values (references 5 in the paper).
-Similar to the stock prices, because BTC is considered a tradable investment instrument, many attempt to predict its price in the future to faciliate their trading strategy.
-Recently studies were published to discuss whether the price of BTC is predictable and technical indicators returned the possibility.
-Accuracy up to 63% for direction prediction (reference here, 9).
-Also, in 10, they reported 98% success rate for daily price forecast.
-Based on recent data 2017 and 2018.
-Both EOD prediction.
-We have even short trading horizons, 1 min, 1 hour.
+Researchers have found that 4,495 literatures related to Bitcoin were published from 2011 to 2020 [@aysan2021ascent].
+More recently, another study stated that more than 9,100 research papers were published on cryptocurrencies in the Web of Science [@bara2024impact].
+In 2015, a journal called Ledger was published by the UNiversity of Pittsburgh's University Library System, and it is the first journal dedicated to cryptocurrencies [@extance2015bitcoin].This reflects the trend of growing interest in digital asset.
+In December, 2019, the were 2,813 different cryptocurrencies exsiting with a 201.47 billion US dollars in their total market capitalization; in particular, bitcoin is the biggest contributor and had a 131.89 billion US dollars market capitalizaiton [@chen2021empirical].
+As I am writing this article on December 4, 2024, the current market capitalization of Bitcoin is 1.95 trillion US dollars, with a 103.22 billion US dollars 24-hour trading volume, accroding to forbes.com [@Forbes].
+As of October 2023, there were a total of 1492 cryptocurrency trading plaforms globally, both centralized and decentralized [@Oh_2024].
+Among all the different cryptocurrencies, Bitcoin is the key player in both the market and in the research field [@aysan2021ascent].
+Therefore, learning more about bitcion can be benificial for both the reserachers, and the public who has interest in cryptocurrency and their tradings.
 
-Income is a fundamental concept as it affects nearly every member of society, influencing access to resources, quality of life, and overall economic stability.
-At the same time, it is also a difficult concept to define, as it can be interpreted differently depending on various perspectives, such as economic, social, legal, and tax purposes.
-Therefore, we must examine various demographical factors that may impact income to understand its concept better.
-Our project defines income as "all forms of net financial resources generated or lost from work, investments, or other activities." This project aims to predict personal income using data from the 2023 American Community Survey (ACS) [@ipums_usa_v15].
-The ACS provides detailed demographic information, including age, education, employment, marital status, mortgage status, veteran status.
-As a result, it provides a robust dataset for studying potential factors that may influence income.
+On the other hand, due to its non-tangible nature, volatility is a key aspect of Bitcoin.
+Much reserach has focused on the price fluctuations of Bitcoin, and compared its volatility to more traditional assets, such as gold.
+For example, [@kurihara2018does] demostrated that Bitcoin's higher volatility compared to gold, indicating its speculative nature.
+In another study, the authors have concluded that Bitcoin's extreme price volatility may interfere with its use as a stable investment instrument.
+As a result, speculative investors are often more attracted to Bitcoin [@chen2021empirical], and looking for profit by capitalizing on its rapid price changes [@baur2018bitcoin].
 
-Due to the timeframe limitations, the findings of this project are specifically applicable to the 2023 dataset.
-As such, the results may vary across different periods, and their generalizability has not been assessed within this project.
-The raw data used in this analysis was sourced from IPUMS using the IPUMS API [@ipums_api].
-We utilized pandas [@pandas] for data processing, and data visualization was carried out using Seaborn [@seaborn].
-Additionally, all tables presented in this report were generated with Plotly [@plotly].
-
-Income inequality is a serious problem and can lead to significant consequences and it may affect social structure, economic stability, as well as policy-making processes.
-Researchers found that demographic factors, such as education level and occupation type, can significantly impact an individual's financial standing [@Autor_2013].
-Similarly, social relationships, such as marital status and household characteristics, affect income dynamics [@WaiteGallagher_2000].
-Therefore, applying machine learning models to analyze the ACS dataset and using the information mentioned to predict personal income is reasonable.
+Despite its extreme volatility, Bitcoin is still widely accepted as a tradable investment vehicle, as many of characteristics are similar to stocks [@chen2021empirical].
+In a very recently published study, the authors developed a model called TradExpert-NM, and it can achieve a 59% accuracy for Bitcoin's daily price movement direction [@ding2024tradexpert].
+In another study, [@mudassir2020time] developed a model that can score up to 65% in accuracy for next-day forecast of Bitcoin's price movement.
+However, many of these "high-performing" models are using various data (social media data in [@ding2024tradexpert] and other technical data [@mudassir2020time]) that are usually too complicated for investors with limited knowledge in machine learning and/or data science.
+Even for models used only more accessible data, such as in [@adcock2019non], they utilized neural network as their model.
 

### 2024-12-05 paper/paper.qmd
EDA graphs saved. Abstract, Intro, Data completed. LFS tracks all parquet files now.

@@
 title: "Comparing Different Machine Learning Models at Predicting Bitcoin Price Moving Direction at Different Time Intervals from 2017 to 2024"
-subtitle: "RNN Can Predict Best at One-Hour Intervals"
+subtitle: "RNN Can Predict Best at One-Hour Intervals and XGB Trains the Fastest"
 author: 
@@
 thanks: "All project related files available at: https://github.com/Jiazhou-Bi/BTC-Price-Prediction"
-date: Nov 20, 2024
+date: Dec 7, 2024
 date-format: long
@@
 
-Cryptocurrencies have drawn much attention of the public recently due to its growing public awareness.
-xxx amount of papers/news about BTC was published (refereces here) from xxxx to xxxx.
-According to xxx (reference here), the value of the market...
-etc.
-The trading volume (references) ...
-It has been adopted by xxx platforms and ....
-(reference) Also, mention the importance of BTC (references needed) and its role of being a indicator of the overall crypto market.
-Therefore, it has become one of the most traded ....
-On the other hand, due to its non-tangible nature, volitility is key aspect.
-xxxx data about its ups and downs.
-Only contains data from 2017 because ....
-fits our scope.
-Regain values (references 5 in the paper).
-Similar to the stock prices, because BTC is considered a tradable investment instrument, many attempt to predict its price in the future to faciliate their trading strategy.
-Recently studies were published to discuss whether the price of BTC is predictable and technical indicators returned the possibility.
-Accuracy up to 63% for direction prediction (reference here, 9).
-Also, in 10, they reported 98% success rate for daily price forecast.
-Based on recent data 2017 and 2018.
-Both EOD prediction.
-We have even short trading horizons, 1 min, 1 hour.
+Cryptocurrencies have drawn much attention from the public recently due to their growing public awareness.
+Researchers have found that 4,495 pieces of literature related to Bitcoin were published from 2011 to 2020 [@aysan2021ascent].
+More recently, another study stated that more than 9,100 research papers were published on cryptocurrencies in the Web of Science [@bara2024impact].
+In 2015, a journal called Ledger was published by the University of Pittsburgh's University Library System, and it is the first journal dedicated to cryptocurrencies [@extance2015bitcoin].
+This reflects the trend of growing interest in digital assets.
+In December 2019, there were 2,813 different cryptocurrencies existing with 201.47 billion US dollars in their total market capitalization; in particular, bitcoin is the most significant contributor and had a 131.89 billion US dollars market capitalization [@chen2021empirical].
+When writing this article on December 4, 2024, the current market capitalization of Bitcoin is 1.95 trillion US dollars, with a 103.22 billion US dollars 24-hour trading volume, according to forbes.com [@Forbes].
+As of October 2023, there were a total of 1492 cryptocurrency trading platforms globally, both centralized and decentralized [@Oh_2024].
+Among all the different cryptocurrencies, Bitcoin is the key player in the market and the research field [@aysan2021ascent].
+Therefore, learning more about Bitcoin can benefit both researchers and the public interested in cryptocurrency and its trade.
 
-Income is a fundamental concept as it affects nearly every member of society, influencing access to resources, quality of life, and overall economic stability.
-At the same time, it is also a difficult concept to define, as it can be interpreted differently depending on various perspectives, such as economic, social, legal, and tax purposes.
-Therefore, we must examine various demographical factors that may impact income to understand its concept better.
-Our project defines income as "all forms of net financial resources generated or lost from work, investments, or other activities." This project aims to predict personal income using data from the 2023 American Community Survey (ACS) [@ipums_usa_v15].
-The ACS provides detailed demographic information, including age, education, employment, marital status, mortgage status, veteran status.
-As a result, it provides a robust dataset for studying potential factors that may influence income.
+On the other hand, volatility is a key aspect of Bitcoin due to its non-tangible nature.
+Much research has focused on the price fluctuations of Bitcoin and compared its volatility to more traditional assets, such as gold.
+For example, [@kurihara2018does] demonstrated that Bitcoin has higher volatility than gold, indicating its speculative nature.
+In another study, the authors have concluded that Bitcoin's extreme price volatility may interfere with its use as a stable investment instrument.
+As a result, speculative investors are often more attracted to Bitcoin [@chen2021empirical] and are looking for profit by capitalizing on its rapid price changes [@baur2018bitcoin].
 
-Due to the timeframe limitations, the findings of this project are specifically applicable to the 2023 dataset.
-As such, the results may vary across different periods, and their generalizability has not been assessed within this project.
-The raw data used in this analysis was sourced from IPUMS using the IPUMS API [@ipums_api].
-We utilized pandas [@pandas] for data processing, and data visualization was carried out using Seaborn [@seaborn].
-Additionally, all tables presented in this report were generated with Plotly [@plotly].
+Despite its extreme volatility, Bitcoin is still widely accepted as a tradable investment vehicle, as many characteristics are similar to stocks [@chen2021empirical].
+In a recently published study, the authors developed a model called TradExpert-NM, which can achieve a 59% accuracy for Bitcoin's daily price movement direction [@ding2024tradexpert].
+In another study, [@mudassir2020time] developed a model that can score up to 65% accuracy for the next-day forecast of Bitcoin's price movement.
+However, many of these "high-performing" models use various data (social media data in [@ding2024tradexpert] and other technical data [@mudassir2020time]) that are usually too complicated for investors with limited knowledge of machine learning and data science.
+Even for models that used only more accessible data, such as in [@adcock2019non], they utilized neural networks as their model.
+However, deep learning models, like deep neural networks, generally require more data than traditional machine learning models and can take longer to train [@janiesch2021machine].
+Therefore, these models pose challenges for individuals who want to start trading Bitcoin with limited knowledge and resources.
+As a result, in this project, three commonly used models for price prediction are compared on their performance and relative training effort by using only commonly available, easy-to-understand features (i.e., High Price, Low Price, Open Price, Close Price, and Trade Volume).
+This project aims to provide some knowledge for beginner crypto-traders who want to trade Bitcoin with some help from machine learning algorithms.
 
-Income inequality is a serious problem and can lead to significant consequences and it may affect social structure, economic stability, as well as policy-making processes.
-Researchers found that demographic factors, such as education level and occupation type, can significantly impact an individual's financial standing [@Autor_2013].
-Similarly, social relationships, such as marital status and household characteristics, affect income dynamics [@WaiteGallagher_2000].
-Therefore, applying machine learning models to analyze the ACS dataset and using the information mentioned to predict personal income is reasonable.
+This project extracted data from Binance using a free Python [@python3] package called ccxt [@ccxt].
+A total of three datasets were extracted.
+The three datasets contain the same time horizon, from August 17, 2017, to November 16, 2024.
+The only difference between them is their time intervals.
+The datasets were extracted at 1-minute, 1-hour, and 1-day intervals.
+Binance was chosen because it is the world's largest cryptocurrency exchange [@kowsmann202176], and it only offers data dating back to August 17, 2017, through its free API.
+Arguably, data before 2017 may be less important for our models' training because the cryptocurrency market was vastly different before 2017, and its price surged significantly in 2017 [@lim2020deep].
+One potential drawback of this approach is that we are relying on the data from Binance as the single point of truth, which may not be the case across the whole time horizon in this case because the quality of the data is not cross-verified with other sources.
 
-We have chosen 2023 ACS data because it offers many demographical details, as well as self-reported total income of the individuals.
-The sample is also representative and covers all geographical locations in the USA.
-Furthermore, it provides easy access to their database through API.
-We use predictive models, including random forest, linear regression, and extreme gradient boosting machine to estimate personal income based on demographic factors to provide insights into which potential factors have the most significant impact on personal income.
-We hope the findings generated from our project can provide insights to policy-making process so that we can, hopefully, reduce income inequality and support further economic development in the future.
+The ultimate goal of this project is to compare and contrast various machine learning models in terms of their technical performance and training effort in predicting Bitcoin's price movement direction at three different time intervals.
+By evaluating models on datasets with 1-minute, 1-hour, and 1-day intervals, this project aims to shed light on the trade-offs between predictive accuracy, computational efficiency, and accessibility for beginner traders.
+Furthermore, individuals with limited expertise in machine learning can leverage relatively simple and accessible data to make more informed, confident trading decisions by learning more about these models.
+In short, this project aims to empower crypto-traders by providing practical guidance on selecting and utilizing machine learning models effectively while balancing ease of use and predictive power.
 
-If we can successfully predict personal income using demographic information from the survey, this analysis will be able to contribute to a nuanced understanding of the American economy in 2023.
-It can help to reveal how demographic changess, educational attainment, and employment status/type can impact personal income.
-On the other hand, if the model cannot predict personal income with a high degree of confidence, this would suggest that the factors included in the analysiscannot not fully capture the intricacies of economic behavior and individual circumstances.
-As a result, additional variables, such as detailed education background like school of graduation, languages spoken, health status, etc, might be necessary in accurately predicting income.
+In this analysis, we explored three different models: Long Short-Term Memory (LSTM), Recurrent Neural Network (RNN), and XGBoost (XGB) to predict Bitcoin's future price movement direction.
+LSTM and RNN are deep learning models that are excellent at handling sequential data.
+Thus, they are commonly used in price-prediction models.
+However, these two models generally need more computational resources, making their training/retraining processes effortful.
+On the other hand, XGB requires less computational effort thanks to its gradient-boosting framework, making it ideal for individuals with limited resources.
+By comparing these models on three different time intervals (1 minute, 1 hour, and 1 day), we aim to evaluate their predictive accuracy, training efficiency, and overall practicality for novice traders.
 
-In this analysis, we explored three different models—Linear Regression, Random Forest, and XGBoost—to predict income based on various demographic factors.
-Each model provides valuable insights into the relationships between income and predictors, such as education, age, industry, and gender.
-Among the models, XGBoost demonstrated the best performance in terms of predictive accuracy, as evidenced by its lower RMSE and superior handling of complex data interactions.
-
-The remainder of this paper is structured as follows: @sec-data introduces the raw dataset and describes the cleaned datasets, along with a preliminary analysis through numerical summaries and visualizations.
-@sec-model describes the three machine learning models being used in our study.
+The remainder of this paper is structured as follows: @sec-data introduces the raw dataset, describes the cleaned datasets, and explains our final datasets with additionally engineered features, along with a preliminary analysis through numerical summaries and visualizations.
+@sec-models describes the three machine learning models being used in our study.
 @sec-results explores key findings from our analysis.
-Lastly, @sec-discussion addresses the limitations of the analysis and offers recommendations for reducing delays and improving subway service reliability.
+Lastly, @sec-discussion addresses the limitations of the analysis and offers recommendations for future research projects.
 
@@
 
-The dataset used here is downloaded from the IPUMS USA website via [@ipums_usa_v15].
-Our project's extracted dataset contains all the responses from the American Community Survey (ACS) from 2023.
-Our extraction included the following variables: total income, dwelling mortgage status, sex of the individual, marital status, education attainment, private or public school, industry or occupation, veteran status, and age.
-Some of the variables (such as educational level, occupation) we have chosen here have a more detailed version that contains more information.
-However, due to the nature of categorical variables, having detailed variables will create more classes of that variable, thus leading to more features for our model when encoding those variables.
-Although more features will improve the overall performance of our predictive model, it will also create issues such as overfitting, needing more computational resources, and lack of interpretability.
-We have chosen higher-level variables in this study for better computational efficiency.
-Furthermore, we have cleaned the data for better data quality.
-Details of our data cleaning processes are provided in @sec-datacleaning.
+As mentioned in @sec-introduction, the datasets were extracted from Binance using a free Python [@python3] package called ccxt [@ccxt].
+The extracted raw datasets contain the following variables: Timestamp, High, Low, Open, Close, and Volume.
+Two new variables were added to the cleaned datasets: direction_t-1 and direction_t+1.
+Lastly, in our final datasets, we have added four more variables: high_diff, low_diff, open_diff, and close_diff.
+Furthermore, some missing data were inputed for better data quality.
+Details of our data cleaning and engineering processes are provided in @sec-datacleaning and @sec-featureengineering.
 In the following subsections, we will review all the variables used in this report and provide some basic descriptive statistics.
@@
 
-### Mortgage
-The Mortgage variable describes the mortgage status of dwelling.
-There are five different values, they are "N/A", "No, owned free and clear", "Check mark on manuscript (probably yes)", "Yes, mortgaged/ deed of trust or similar debt", and "Yes, contract to purchase".
-We have dropped the N/A values for better data quality.
-Details available in @sec-datacleaning.
-The count of the remaining classes is shown below in (@fig-figure1).
+### Timestamp
+The timestamp variable in our datasets is a key component for tracking the temporal aspect of Bitcoin's price movement.
+All three datasets start on August 17, 2017, and end on November 16, 2024.
+While ensuring all three datasets cover the same period, each dataset has different granularities.
+In the 1-minute dataset, the timestamp is recorded for each minute, providing the highest resolution data.
+In the 1-hour dataset, the timestamp is aggregated at hourly intervals.
+In the 1-day dataset, the timestamps are recorded at daily intervals.
+The differences in temporal granularity allow us to examine the trade-offs between datasets containing more information closely but may also increase computational complexity and noise.
+@tbl-table1 is an example of the timestamp variable from the 1-hour dataset.
+
+```{python}
+
+#| echo: false
+#| label: tbl-table1
+#| tbl-cap: "Example of the timestamp variable from the 1-hour dataset."
+#| warning: false
+#| message: false
+#| fig-pos: "H"
+
+import pandas as pd
+data = pd.read_parquet('../data/01-raw_data/raw_data_1h.parquet')
+df = data.drop(columns=['open', 'high', 'low', 'close', 'volume'])
+df = df.set_index('timestamp')
+
+df.head()
+
+```
+
+### Open
+Open refers to the 'open price' of the tradable asset.
+As defined in [@chen2021empirical], the open price is the initial price of the cryptocurrency at the beginning of the trading session.
+For example, in the 1-hour dataset, the open price is the price of Bitcoin at the start of the hour.
+This is a commonly used indicator in the financial sector, and it shows the perceived value change of the underlying asset over time.
+@fig-figure1 reveals the change of the 1-hour open price of Bitcoin over the entire time horizon of the dataset.
 
@@
 #| label: fig-figure1
+#| fig-cap: "Line chart of the open prices from the 1-hour dataset."
 #| warning: false
@@
 #| fig-pos: "H"
-#| fig-cap: Most Homeowners Are on Mortgage
 
@@
 import matplotlib.pyplot as plt
-import seaborn as sns
-data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
-mortgage_labels = {
-    '1': 'No, owned free and clear',
-    '3': 'Yes, mortgaged',
-    '4': 'Yes, contract to purchase'
-}
-data['MORTGAGE'] = data['MORTGAGE'].astype(str).map(mortgage_labels)
-plt.figure(figsize=(5, 3))
-sns.countplot(x='MORTGAGE', data=data, palette='pastel', edgecolor='black')
-plt.xlabel('Mortgage Status', fontsize=12)
-plt.ylabel('Count', fontsize=12)
-plt.xticks(rotation=30, ha='right')
-plt.grid(axis='y', linestyle='--', alpha=0.7)
+data = pd.read_parquet('../data/01-raw_data/raw_data_1h.parquet')
+
+plt.figure(figsize=(10, 6))
+plt.plot(data['timestamp'], data['open'], linewidth=0.5)
+plt.title('Line Chart of Open Prices (1-Hour Interval)', fontsize=14)
+plt.xlabel('Timestamp')
+plt.ylabel('Open Price (USD)')
+plt.ylim(0, None)
+plt.grid(True)
 plt.tight_layout()
@@
 
-### Sex
-This variable is coded as a binary variable, and we will use it as-is in our project.
-This does not suggest that the authors agree with this type of sex classification, but rather that it aligns with the available data and serves the research purposes of this project.
-The binary coding is a practical choice to maintain consistency with the dataset, though it may not fully capture the complexity and diversity of gender identities.
-@tbl-table1 shows the distribution of this variable.
+### Close {#sec-close}
+Similar to 'open,' 'close' refers to the 'close price' of the tradable asset.
+As defined in [@chen2021empirical], the close price is the last recorded price of Bitcoin before the trading session ends.
+For example, in the 1-minute dataset, the open price is the price of Bitcoin at the end of the minute.
+This variable is widely used in financial analysis as a key indicator of market sentiment and performance over a given time period.
+More importantly, close price is often considered the most robust indicator of the assets' perceived price, as suggested in [@Hayes].
+Therefore, 'close' is also used as the 'final price' of Bitcoin in this project.
+@fig-figure2 shows the change of the 1-minute open price of Bitcoin over the entire time horizon of the dataset.
 
@@
 #| echo: false
-#| label: tbl-table1
+#| label: fig-figure2
+#| fig-cap: "Line chart of the close prices from the 1-minute dataset."
 #| warning: false
@@
 
-data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
-sex_labels = {'1': 'Male', '2': 'Female'}
-data['SEX'] = data['SEX'].map(sex_labels)
-vet_status_count = data['SEX'].value_counts().reset_index()
-vet_status_count.columns = ['Sex', 'Count']
-fig, ax = plt.subplots(figsize=(6, 2))
-ax.axis('off')
-table = ax.table(
-    cellText=vet_status_count.values,
-    colLabels=vet_status_count.columns,
-    cellLoc='center',
-    loc='center'
-)
-table.auto_set_font_size(False)
-table.set_fontsize(10)
-table.scale(1, 1.5)
+data = pd.read_parquet('../data/01-raw_data/raw_data_1m.parquet')
+
+# Plot the 'close' variable
+plt.figure(figsize=(10, 6))
+plt.plot(data['timestamp'], data['close'], linewidth=0.5)
+plt.title('Line Chart of Close Prices (1-Minute Interval)', fontsize=14)
+plt.xlabel('Timestamp')
+plt.ylabel('Close Price (USD)')
+plt.ylim(0, None)
+plt.grid(True)
 plt.tight_layout()
@@
 
-### Age
-Age describes the age of the individual, and a age of 0 was coded for the individuals under 1 year old.
-For our dataset, the minimum age is 16, however.
-Age distribution of our dataset is shown below in @fig-figure2.
-```{python}
+### High and Low {#sec-highandlow}
+The high and low variables represent the extremes of Bitcoin's price within a given time interval.
+As suggested by the names, high refers to the highest price of Bitcoin during the time interval, and low refers to the lowest.
+[@lapitskaya2024prediction], [@poudel2023cryptocurrency], and many other researchers believe that incorporating basic features such as high and low prices can significantly improve the performance of cryptocurrency price prediction models.
+In @fig-figure3, both high and low from the 1-hour dataset are presented in the same graph.
+It is clear that these two are highly correlated, and this problem will be addressed later in @sec-featureengineering.
 
-#| echo: false
-#| label: fig-figure2
-#| warning: false
-#| message: false
-#| fig-pos: "H"
-#| fig-cap: Age Distribution
-
-import pandas as pd
-import matplotlib.pyplot as plt
-import seaborn as sns
-data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
-plt.figure(figsize=(10, 6))
-sns.countplot(x='AGE', data=data, palette='pastel', edgecolor='black')
-plt.xlabel('Age', fontsize=12)
-plt.ylabel('Count', fontsize=8)
-plt.xticks(rotation=90, ha='center')
-plt.grid(axis='y', linestyle='--', alpha=0.7)
-plt.tight_layout()
-plt.show()
-
-```
-
-### Marital Status
-Marital status describes the marital status of the individual.
-There are 7 different values, namely "Married, spouse present", "Married, spouse absent", "Separated", "Divorced", "Widowed", "Never married/single", and "Blank, missing".
-We have dropped the N/A values for better data quality.
-Details available in @sec-datacleaning.
-The following @fig-figure3 illustrate the distribution of marital status of the participants in our dataset.
 ```{python}
@@
 #| label: fig-figure3
-#| warning: false
-#| message: false
-#| fig-pos: "H"
-#| fig-cap: Counts of Marital Status, Most Participants Are Married with Spouse Present, or Never Married/Single
-
-import pandas as pd
-import matplotlib.pyplot as plt
-import seaborn as sns
-data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
-marrital_status_label = {
-  '1': 'Married, spouse present',
-  '2': 'Married, spouse absent',
-  '3': 'Separated',
-  '4': 'Divorced',
-  '5': 'Widowed',
-  '6': 'Never married/single',
-  '9': 'Blank, missing'
-}
-order = [
-    'Married, spouse present',
-    'Married, spouse absent',
-    'Separated',
-    'Divorced',
-    'Widowed',
-    'Never married/single'
-]
-data['MARST'] = data['MARST'].map(marrital_status_label)
-plt.figure(figsize=(10, 6))
-sns.countplot(x='MARST', data=data, order=order, palette='pastel', edgecolor='black')
-plt.xlabel('Marital Status', fontsize=12)
-plt.ylabel('Count', fontsize=12)
-plt.xticks(rotation=30, ha='center')
-plt.grid(axis='y', linestyle='--', alpha=0.7)
-plt.tight_layout()
-plt.show()
-
-```
-
-### Educational Attainment
-This is a general version (compared to the detailed version).
-The reason of choosing the general version is mentioned in @sec-data above.
-This column represents the highest educational attainment achieved by the individual.
-The distribution of the variable is shown in @fig-figure4.
-```{python}
-
-#| echo: false
-#| label: fig-figure4
-#| warning: false
-#| message: false
-#| fig-pos: "H"
-#| fig-cap: Most Participants Have Middel School Attainment or Above
-
-import pandas as pd
-import matplotlib.pyplot as plt
-import seaborn as sns
-data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
-educ_new_label = {
-    0: 'No Schooling',
-    1: 'Nursery School',
-    2: 'Primary School',
-    3: 'Middle School',
-    4: 'College 1-4 Years',
-    5: 'College 5+ Years'
-}
-order = [
-    'No Schooling',
-    'Nursery School',
-    'Primary School',
-    'Middle School',
-    'College 1-4 Years',
-    'College 5+ Years'
-]
-data['EDUC_new'] = data['EDUC_new'].map(educ_new_label)
-plt.figure(figsize=(10, 6))
-sns.countplot(x='EDUC_new', data=data, order=order, palette='pastel', edgecolor='black')
-plt.xlabel('Educational Attainment', fontsize=12)
-plt.ylabel('Count', fontsize=12)
-plt.xticks(rotation=30, ha='center')
-plt.grid(axis='y', linestyle='--', alpha=0.7)
-plt.tight_layout()
-plt.show()
-
-```
-
-### School Type
-School type indicates if the individual went to a public school or private school.
-Other than missing values, there are three different classes.
-They are "Not enrolled", "Public school", and "Private school".
-We are not sure how the participants would respond if they have been to more than one type of schools.
-@fig-figure5 shows the distribution of our sample.
-```{python}
-
-#| echo: false
-#| label: fig-figure5
-#| warning: false
-#| message: false
-#| fig-pos: "H"
-#| fig-cap: School Type
-
-import pandas as pd
-import matplotlib.pyplot as plt
-import seaborn as sns
-data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
-school_type_label = {
-    "1": 'Not Enrolled',
-    "2": 'Private School',
-    "3": 'Public School',
-}
-data['SCHLTYPE'] = data['SCHLTYPE'].map(school_type_label)
-plt.figure(figsize=(10, 6))
-sns.countplot(x='SCHLTYPE', data=data, palette='pastel', edgecolor='black')
-plt.xlabel('Private or Public School', fontsize=12)
-plt.ylabel('Count', fontsize=12)
-plt.xticks(rotation=30, ha='center')
-plt.grid(axis='y', linestyle='--', alpha=0.7)
-plt.tight_layout()
-plt.show()
-
-```
-
-### Industry
-Industry describes the type of industry the individual works in.
-Same as the educational attainment variable, more detailed version of this variable is available (i.e., Occupation).
-There are more than 200 industries listed in our sample.
-A detailed description can be seen at [IPUMS USA](https://usa.ipums.org/usa-action/variables/IND1990#description_section).
-
-### Veteran Status
-Veteran status idicates whether the individual is a veteran or not.
-It can be either "Yes", "No", or "N/A".
-Please notice that "N/A" here is different from missing, suggesting this might be a different indicator, which is not explicitly explained from the ddl file provided.
-There is also a detailed version of this variable, which is not used in our project.
-The counts is displayed in @tbl-table2.
-```{python}
-
-#| echo: false
-#| label: tbl-table2
+#| fig-cap: "Line chart of high and low prices from the 1-hour dataset."
 #| warning: false
@@
 
-data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
-vet_status_labels = {'0': 'N/A', '1': 'No', '2': 'Yes'}
-data['VETSTAT'] = data['VETSTAT'].map(vet_status_labels)
-vet_status_count = data['VETSTAT'].value_counts().reset_index()
-vet_status_count.columns = ['Veteran Status', 'Count']
-fig, ax = plt.subplots(figsize=(6, 2))
-ax.axis('off')
-table = ax.table(
-    cellText=vet_status_count.values,
-    colLabels=vet_status_count.columns,
-    cellLoc='center',
-    loc='center'
-)
-table.auto_set_font_size(False)
-table.set_fontsize(10)
-table.scale(1, 1.5)
-plt.tight_layout()
-plt.show()
-```
-
-### Total Income
-Total income is the total income made by the individual for the year.
-The distribution of total income is depicted in @fig-figure6.
-We have removed the outliers from the graph to have a better understanding of the general pattern of the overall distribution.
-From the figure, it is clear that more income are reported near multiple of $10,000 as a unit.
-This may be caused by the "heaping and leaping" phenomonon found in self-reported survey data [@pudney2008heaping].
-```{python}
-
-#| echo: false
-#| label: fig-figure6
-#| warning: false
-#| message: false
-#| fig-pos: "H"
-#| fig-cap: Total Income Distribution
-
-import pandas as pd
-import matplotlib.pyplot as plt
-import seaborn as sns
-data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
+data = pd.read_parquet('../data/01-raw_data/raw_data_1h.parquet')
 plt.figure(figsize=(10, 6))
-sns.histplot(x='INCTOT', data=data, palette='pastel', edgecolor='black')
-plt.xlabel('Total Income', fontsize=12)
-plt.ylabel('Count', fontsize=12)
-plt.xticks(rotation=30, ha='center')
-plt.xlim(0, 250000)
-plt.grid(axis='y', linestyle='--', alpha=0.7)
+plt.plot(data['timestamp'], data['high'], label='High Price', linewidth=0.5)
+plt.plot(data['timestamp'], data['low'], label='Low Price', linewidth=0.5)
+plt.title('Line Chart of High and Low Prices (1-Hour Interval)', fontsize=14)
+plt.xlabel('Timestamp')
+plt.ylabel('Price (USD)')
+plt.ylim(0, None)
+plt.legend()
+plt.grid(True)
 plt.tight_layout()
@@
 
-## Analysis of Feature Relationships {#sec-afr}
+### Volume
+The last raw variable extracted is volume.
+Volume represents the total amount of Bitcoin traded during a given time interval and indicates market activity and liquidity.
+For example, in the 1-hour dataset, the volume indicates the cumulative number of Bitcoins traded within an hour.
+In [@osina2023exploring], the authors have found that trade volume is an important feature when predicting the price of three different cryptocurrencies.
+Generally speaking, higher trading volume reflects increased market interest and is often associated with significant price fluctuations.
+As shown in @fig-figure4, the trading volume in the 1-day dataset varies significantly over time, illustrating periods of heightened trading activity.
 
-In this subsection, we will explore some of the potential relationships among the features chosen for our study.
-Understanding these relationships can prepare us better for our model construction in the next section.
+```{python}
 
-### Income Distribution by Mortgage Status
+#| echo: false
+#| label: fig-figure4
+#| fig-cap: "Line chart of trading volume from the 1-day dataset."
+#| warning: false
+#| message: false
+#| fig-pos: "H"
 
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.8\textwidth]{../other/graphs/EDAincomebymortgage.png}
-    \caption{Income Distribution by Mortgage Status}
-    \label{fig:income-mortgage}
-\end{figure}
+import pandas as pd
+import matplotlib.pyplot as plt
 
-Figure \ref{fig:income-mortgage} illustrates the income distribution for individuals with different mortgage statuses.
-The highest average income is observed among individuals with a "Mortgage/Trust Deed," followed by those who own their homes outright ("Fully Owned").
-Those with a "Signed Purchase Contract" have the lowest average income among the groups.
-This pattern may reflect a correlation between mortgage status and income level, where individuals with higher incomes are more likely to secure traditional mortgage loans, while those with lower incomes tend towards alternative purchasing arrangements.
-This could indicate the influence of financial stability and access to credit on mortgage status.
+# Load the dataset
+data = pd.read_parquet('../data/01-raw_data/raw_data_1d.parquet')
 
-### Correlation Analysis of Age, Education, and Income
+# Plot the 'volume' variable
+plt.figure(figsize=(10, 6))
+plt.plot(data['timestamp'], data['volume'], linewidth=0.5, color='orange')
+plt.title('Line Chart of Trading Volume (1-Day Interval)', fontsize=14)
+plt.xlabel('Timestamp')
+plt.ylabel('Trading Volume (BTC)')
+plt.ylim(0, None)
+plt.grid(True)
+plt.tight_layout()
+plt.show()
 
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.8\textwidth]{../other/graphs/EDAcorrelation.png}
-    \caption{Correlation Matrix of Age, Education, and Income}
-    \label{fig:correlation-matrix}
-\end{figure}
+```
 
-Figure \ref{fig:correlation-matrix} shows the correlation matrix for three key variables: age, education level (EDUC_new), and total personal income (INCTOT).
-The matrix reveals that there is a moderate positive correlation between education level and income (0.29), suggesting that higher education levels are generally associated with higher income.
-Additionally, a smaller positive correlation exists between age and income (0.14), which might indicate that as individuals grow older, they tend to earn more due to accumulated experience.
-The correlation between age and education (0.06) is minimal, which is expected as these two variables are less directly related.
-Overall, this analysis supports the notion that education level plays a more significant role in income variation than age within this dataset.
+## Data Cleaning {#sec-datacleaning}
+There was no "obvious" missing value from the extracted datasets, as all the mentioned variables above are filled with either a timestamp in the "timestamp" column or a numeric value in all the remaining columns.
+However, upon a closer examination, the timestamps were not exhaustive.
+For example, in the 1-minute dataset, the timestamp "2017-09-06 16:01:00" does not exist even though it is included within the time horizon of the dataset.
+8632 and 128 rows were missing from the 1-minute and 1-hour datasets, respectively.
+As a result, linear interpolation was used to estimate the missing values because it is a straightforward technique to fill in missing values in time series data [@moritz2015comparison].
+Meanwhile, a new column was added to the 1-minute and 1-hour datasets to indicate if the prices of the row were either existing from the raw dataset or input.
+The new column is named "was_missing." The value of "was_missing" is 0 if the prices were extracted from the original data and 1 if the prices were filled with linear interpolation.
+Other than interpolating the prices of missing timestamps, no other data were manipulated.
 
-### Average Income by Education and Gender {#sec-afr3}
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.8\textwidth]{../other/graphs/EDAinteraction.png}
-    \caption{Average Income by Education and Sex}
-    \label{fig:income-education-sex}
-\end{figure}
+## Feature Engineering {#sec-featureengineering}
+First, two new features were added to indicate the price-moving direction of Bitcoin, namely "direction_t-1" and "direction_t+1".
+These features provide insights into the direction of Bitcoin's price change relative to the current timestamp.
+Second, four more new features were engineered to capture Bitcoin's price movements: "open_diff," "high_diff," "low_diff," and "close_diff." These features represent the lagged differences of the respective price variables (open, high, low, and close) from one timestamp to the next.
 
-Figure \ref{fig:income-education-sex} presents the average total personal income based on different levels of education for both male and female groups.
-From the plot, it is observed that individuals with higher levels of education, particularly those who attended "College 5+ Years," tend to earn more on average compared to other groups.
-However, there are some variations in income within each education level, as indicated by the error bars.
-Generally, males tend to have higher average incomes across most education levels.
-This difference suggests a possible gender-based income disparity, even after controlling for education levels.
-Additionally, the variability in income among education levels highlights that factors beyond educational attainment, such as occupation and industry, may further impact income.
-Thus, incorporating an interaction term between education and gender in the model is necessary to capture these nuanced effects more accurately.
+### Direction_t-1
+This feature indicates the price movement direction compared to the previous timestamp.
+It was calculated by calculating the difference between the close price of the current row and the previous row.
+If the price has decreased from the previous timestamp, "direction_t-1" is marked as -1; if the close price has increased, then "direction_t-1" is marked as 1; if the close price did not change, it is assigned 0—the reason for choosing the close price as the final price indicator was explained in @sec-close.
 
-# Model {#sec-model}
+### Direction_t+1
+Similar to 'direction_t-1', 'direction_t+1' indicates the price movement direction at the next timestamp relative to the current close price, and it was calculated by calculating the close difference from the next timestamp to the current close price.
+"direction_t+1" is marked as -1 if the close price is lower in the next timestamp, and is marked as 1 if the close price is going up; if the close price does not change, it is assigned 0—the reason for choosing the close price as the final price indicator was explained in @sec-close.
+Because 'direction_t+1' indicates the price movement direction of the current timestamp, this variable is used as our models' outcome (i.e., the result the models try to predict).
 
-Our modeling strategy aims to explore the relationship between personal characteristics and income.
-We implemented three models: Linear Regression with Interaction Terms, Random Forest, and XGBoost, each chosen for their unique strengths in prediction and interpretation.
+### Open_Diff, High_Diff, Low_Diff, and Close_Diff
+An issue was mentioned in @sec-highandlow.
+That is, the correlation between the price variables (open, high, low, and close) is too high (almost or equal to 1), as illustrated below in @fig-figure5 and @fig-figure6.
 
-## Train Test Split of Data
+```{python}
 
-We have cleaned our data before we split it into training and testing datasets.
-We are aware of potential data leakage of our testing data, and we have mitigated this risk by not using overall statistics or normalization that requires the original data to calculate, not imputing any missing data, or engineering features that include the original dataset.
-By doing so, we can minimize the risk of data leakage, and ensure the perfect seventy and thirty percent train test split.
+#| echo: false
+#| label: fig-figure5
+#| fig-cap: "All price variables (open, high, low, close) plotted for the 1-minute dataset."
+#| warning: false
+#| message: false
+#| fig-pos: "H"
 
-## Linear Regression Model with Interaction Term {#sec-linear-regression-model-with-interaction-term}
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
 
-### Model set-up
+img = mpimg.imread('../other/graphs/all_price_plot_1m.png')
+plt.figure(figsize=(10, 6))
+plt.imshow(img)
+plt.axis('off')
+plt.show()
 
-We implemented a linear regression model with an interaction term to explore the relationship between personal characteristics and income. The general form of the model is:
+```
 
-$$
-y=\beta_0+\beta_1X_1+\beta_2X_2+...+\beta_nX_n+\beta_{\text{interaction}}(X_i\times X_j)+\epsilon
-$$
+```{python}
 
-where:
+#| echo: false
+#| label: fig-figure6
+#| fig-cap: "Correlation matrix (open, high, low, close) plotted for the 1-minute dataset."
+#| warning: false
+#| message: false
+#| fig-pos: "H"
 
--   y represents the total income (`INCTOT`),
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
 
--   $X_1, X_2, ... , X_n$ are the predictor variables, including both categorical and numerical features,
+img = mpimg.imread('../other/graphs/all_price_correlation_1m.png')
+plt.figure(figsize=(10, 6))
+plt.imshow(img)
+plt.axis('off')
+plt.show()
 
--   $X_i \times X_j$ is the interaction term between education and sex (`EDUC_SEX_INTERACTION`)
+```
 
--   $\beta_0$ is the intercept,
+To solve this issue, "open_diff," "high_diff," "low_diff," and "close_diff" were added to include the lagged differences in the prices.
+Lagged difference refers to the difference in prices from the current price to the price from the previous timestamp.
+They are often used in predictive models in financial time series to solve the multicollinearity issue with the original prices, as suggested in [@moews2019lagged].
+In @fig-figure7 and @fig-figure8 below, the correlation between the lagged difference price variables (open_diff, high_diff, low_diff, and close_diff) were displayed.
 
--   $\beta_1, \beta_2, ... , \beta_n, \beta_{\text{interaction}}$ are the coefficients for the respective predictors and interaction term.
+```{python}
 
-To ensure the robustness of our linear regression model, we conducted a series of steps aimed at minimizing multicollinearity and improving model interpretability.
-Initially, we included all relevant predictor variables, both categorical and numerical, to explore the full range of relationships between personal characteristics and income.
-Next, we calculated the Variance Inflation Factor (VIF) for each predictor to identify potential multicollinearity issues.
-Typically, a VIF value greater than 10 indicates problematic multicollinearity, which can distort the model's estimates.
-In our analysis, variables such as `VESTAT` (Veteran Status) and `AGE` had VIF values exceeding 13, suggesting they were highly correlated with other predictors.
-To address this, we removed these high-VIF variables from the model.
-Afterward, we refit the linear regression model using the reduced set of predictors, which resulted in improved performance metrics, including a lower Mean Squared Error and a more interpretable set of coefficients.
+#| echo: false
+#| label: fig-figure7
+#| fig-cap: "All lagged price differences (open_diff, high_diff, low_diff, and close_diff) plotted for the 1-minute dataset."
+#| warning: false
+#| message: false
+#| fig-pos: "H"
 
-Note that according to our previous analysis in @sec-afr3, the interaction term, `EDUC_SEX_INTERACTION`, was included to capture the combined effect of education level and gender on income.
-This allowed us to model the non-linear relationship between these two variables and their joint impact on the outcome variable.
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
 
-::: {#tbl-coefficients}
-\begin{center}
-Table 2: Coefficients from the Linear Regression Model
-\end{center}
+img = mpimg.imread('../other/graphs/all_price_diff_plot_1m.png')
+plt.figure(figsize=(10, 6))
+plt.imshow(img)
+plt.axis('off')
+plt.show()
 
-| Feature                     | Coefficient |
-|-----------------------------|-------------|
-| cat__MORTGAGE             | 11267.90    |
-| num__AGE                  | 415.91      |
-| cat__STATEICP             | 126.39      |
-| cat__IND1990              | -43.18      |
-| cat__OCC2010              | -139.55     |
-| num__EDUC_SEX_INTERACTION | -1832.32    |
-| cat__MARST                | -5541.39    |
-| cat__GQ                   | -7956.89    |
-| cat__SCHLTYPE             | -16320.74   |
-:::
+```
 
-*Table 2: Linear Regression Coefficients Table.*
+```{python}
 
-We run the model in Python [@cite_python] using the `scikit-learn` package [@cite_sklearn].
-We use the default settings for the `LinearRegression` function from `scikit-learn`.
-We use the `statsmodels` package [@cite_statsmodels] to calculate the VIF for each predictor.
+#| echo: false
+#| label: fig-figure8
+#| fig-cap: "Correlation matrix (open_diff, high_diff, low_diff, and close_diff) plotted for the 1-minute dataset."
+#| warning: false
+#| message: false
+#| fig-pos: "H"
 
-### Model Justification
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
 
-In evaluating our model's performance, we selected key metrics to assess both its fit and predictive accuracy, including the Variance Inflation Factor (VIF), R-squared ($R^2$), and Mean Squared Error (MSE).
-These metrics were chosen to provide a comprehensive evaluation of the model's robustness and reliability in capturing the relationship between personal characteristics and income.
+img = mpimg.imread('../other/graphs/all_price_diff_correlation_1m.png')
+plt.figure(figsize=(10, 6))
+plt.imshow(img)
+plt.axis('off')
+plt.show()
 
-The VIF was used to diagnose multicollinearity among predictors, helping us ensure that no variables were highly correlated, which could distort the interpretation of the model's coefficients.
+```
 
-::: {#tbl-VIF}
-\begin{center}
-Table 3: Variance Inflation Factors (VIF) for Predictor Variables
-\end{center}
-
-| Feature                  | VIF       |
-|--------------------------|-----------|
-| cat__STATEICP            | 3.24  |
-| cat__GQ                  | 1.00  |
-| cat__MORTGAGE            | 2.66  |
-| cat__MARST               | 1.74  |
-| cat__SCHLTYPE            | 1.24  |
-| cat__OCC2010             | 3.19  |
-| cat__IND1990             | 8.49  |
-| num__AGE                 | 7.87  |
-| num__EDUC_SEX_INTERACTION| 6.95  |
-:::
-
-*Table 3: This table displays the Variance Inflation Factors (VIF) for each of the predictor variables in the model. VIF values greater than 10 indicate potential multicollinearity issues.*
-
-We chose $R^2$ as it provides a measure of how well the model explains the variance in the dependent variable, making it an important indicator of the model’s explanatory power.
-The MSE was selected to quantify the average squared difference between observed and predicted values, offering a clear view of the model’s prediction error.
-
-::: {#tbl-modelperformance}
-\begin{center}
-Table 4: Model Performance Metrics
-\end{center}
-
-| Metric                     | Value              |
-|----------------------------|--------------------|
-| R-squared                  | 0.0774       |
-| Adjusted R-squared         | 0.0774     |
-| Mean Squared Error (MSE)   | 7,448,800,507.01   |
-| Root Mean Squared Error (RMSE) | 86,306.43        |
-:::
-
-*Table 4: This table presents the performance metrics of the linear regression model, including MSE, RMSE, R-squared, and Adjusted R-squared.
-These metrics are used to assess the model's fit and predictive accuracy.*
-
-Additionally, residual analysis was conducted by plotting residuals against fitted values to check for randomness around zero, ensuring that the model satisfied the assumptions of linearity and homoscedasticity.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=0.5\textwidth]{../other/graphs/LinearModelResidual1.png}
-\caption{Linear Model Residuals Plot}
-\label{fig:residual-linear1}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=0.5\textwidth]{../other/graphs/LinearModelResidual2.png}
-\caption{Linear Model Residuals vs. Predicted Values Plot}
-\label{fig:residual-linear2}
-\end{figure}
-
-We use the statsmodels package [@cite_statsmodels] to calculate the VIF for each predictor.
-We use the matplotlib [@cite_matplotlib] and seaborn [@seaborn] to visualize the residuals.
-We use the scikit-learn [@cite_sklearn] package to calculate the $R^2$ score.
-
-## Random Forest Model {#sec-random-forest}
-
-### Model set-up
-
-To analyze the non-linear relationship between personal characteristics and income, we employed a Random Forest model.
-Random Forest is an ensemble learning method that builds multiple decision trees and averages their predictions, reducing overfitting and capturing complex, non-linear relationships in the data.
-We utilized Bagging (Bootstrap Aggregating) and random feature selection to enhance model performance and reduce overfitting.
-Bagging involves generating multiple bootstrap samples from the original training set, with each sample used to train an independent decision tree.
-By averaging the predictions of multiple trees, Bagging reduces the variance of the model, thereby improving its robustness.
-Additionally, Random Forest introduces further randomness by selecting a subset of features at each split, which reduces correlation among trees and improves the model’s generalization ability.
-
-To optimize the model, we conducted hyperparameter tuning, focusing on two key parameters: `n_estimators` (the number of trees) and `max_depth` (the maximum depth of each tree).
-Increasing `n_estimators` typically decreases model variance and improves stability, but too many trees can result in diminishing returns with increased computation time.
-Initially, we set `n_estimators` to 100, but through Grid Search, we tested values of 100, 200, and 300 and found that 300 trees yielded the best performance.
-Similarly, the `max_depth` parameter controls the complexity of the trees.
-While deeper trees can capture more intricate data patterns, they may also overfit.
-We initially set `max_depth` to 15 and test 10, 15, 20 but found that reducing it to 10 during hyperparameter tuning improved generalization by avoiding overfitting.
-Additionally, we tuned `min_samples_split` and `min_samples_leaf` to prevent overfitting by ensuring that nodes have a sufficient number of samples before splitting.
-
-We run the model in Python [@cite_python] using the scikit-learn package [@cite_sklearn] and pandas [@pandas].
-We used the GridSearchCV function from scikit-learn for hyperparameter tuning and subsampled the training data using pandas.
-The RandomForestRegressor function from scikit-learn was employed for model fitting.
-
-### Model Justification
-
-To assess the performance of our Random Forest model, we evaluated several key metrics, including Mean Squared Error (MSE), Root Mean Squared Error (RMSE), feature importance, and residual analysis.
-These evaluations provided insights into the model’s predictive accuracy, interpretability, and overall fit.
-
-First, we calculated the Mean Squared Error (MSE) and Root Mean Squared Error (RMSE) to assess the overall prediction accuracy of the model.
-These metrics provided insight into how well the model captured the variance in income based on the selected features.
-
-::: {#tbl-randomforestMSE}
-\begin{center}
-Table 5: Best Random Forest Model Performance Metrics
-\end{center}
-
-| Metric          | Value            |
-|-----------------|------------------|
-| Best Model MSE  | 6,057,426,411.65 |
-| Best Model RMSE | 77,829.47        |
-:::
-
-*Table 5: This table shows the Mean Squared Error (MSE) and Root Mean Squared Error (RMSE) for the best-tuned Random Forest model, demonstrating the model's performance after hyperparameter tuning.*
-
-Additionally, we performed a feature importance analysis, which allowed us to understand the relative significance of each predictor variable in determining income.
-This is particularly useful for verifying that the model identifies the key factors influencing income as expected.
-
-::: {#tbl-randomforestimportance}
-\begin{center}
-Table 6: Feature Importance in the Best Random Forest Model
-\end{center}
-
-| Feature         | Importance |
-|-----------------|------------|
-| cat\_\_EDUC_new | 0.337   |
-| cat\_\_IND1990  | 0.243   |
-| num\_\_AGE      | 0.194   |
-| cat\_\_SEX      | 0.149   |
-| cat\_\_MARST    | 0.056   |
-| cat\_\_MORTGAGE | 0.013   |
-| cat\_\_SCHLTYPE | 0.004   |
-| cat\_\_VETSTAT  | 0.003   |
-:::
-
-*Table 6: This table lists the feature importance values for the best-performing Random Forest model.
-The higher the importance value, the more significant the feature is in predicting the target variable (income).*
-
-We also conducted a residual analysis by plotting the residuals (the difference between actual and predicted values) against the predicted values.
-This step helped us evaluate whether the model appropriately captured the underlying patterns in the data, ensuring that there were no significant biases or violations of model assumptions.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=0.5\textwidth]{../other/graphs/RandomForestResidual.png}
-\caption{Random Forest Model Residuals vs. Predicted Values Plot}
-\label{fig:residual-randomforest}
-\end{figure}
-
-The feature importances were extracted from the best model using RandomForestRegressor from scikit-learn [@cite_sklearn] and displayed using pandas [@pandas].
-We use the matplotlib [@cite_matplotlib] to visualize the residuals.
-
-## XGBoost {#sec-xgboost}
-
-### Model set-up
-
-XGBoost (Extreme Gradient Boosting) is a powerful machine learning algorithm that builds an ensemble of decision trees using gradient boosting techniques.
-In gradient boosting, each subsequent tree is trained to correct the residuals of the previous trees, iteratively improving the model’s predictive accuracy.
-XGBoost enhances this process by incorporating regularization techniques to control overfitting, making it an ideal choice for capturing complex, non-linear relationships in data.
-
-To optimize the performance of our XGBoost model, we employed GridSearchCV to conduct hyperparameter tuning.
-We explored a range of hyperparameters to find the optimal configuration for our dataset.
-The grid search included the learning rate (`learning_rata`), max depth (`max_depth`), number of trees (`n_estimators`), and minimum child weight (`min_child_weight`).
-Specifically, we tested learning rates of 0.01, 0.05, and 0.1, tree depths of 4, 6, and 8, and `n_estimators` values of 100, 200, and 300.
-Using GridSearchCV with a cross-validation fold of 3, we systematically evaluated different combinations of these hyperparameters.
-The model was evaluated using negative Mean Squared Error (MSE) as the scoring metric.
-The optimal model used a learning rate of `0.1`, a maximum tree depth of `6`, a minimum child weight of `5`, and `300` boosting rounds.
-This configuration provided a good balance between model complexity and predictive accuracy, capturing important non-linear relationships while minimizing overfitting.
-
-We implemented the XGBoost model in Python [@cite_python] using the scikit-learn package [@cite_sklearn], with pandas [@pandas] for data processing and xgboost [@cite_xgboost] for model fitting.
-The initial model was trained using default parameters of the XGBRegressor function, which was subsequently fine-tuned using GridSearchCV from scikit-learn [@cite_sklearn] for hyperparameter optimization.
-
-### Model Justification
-
-Similar to the Random Forest model, the XGBoost model was trained on a 70-30 train-test split.
-We evaluated the model's performance using Mean Squared Error (MSE) and Root Mean Squared Error (RMSE) to assess its predictive accuracy.
-
-::: {#tbl-XGBoostMSE}
-\begin{center}
-Table 7: XGBoost Model Performance Metrics
-\end{center}
-
-| Metric                         | Value            |
-|--------------------------------|------------------|
-| Mean Squared Error (MSE)       | 5,854,629,848.13 |
-| Root Mean Squared Error (RMSE) | 76,515.55        |
-:::
-
-*Table 7: This table shows the Mean Squared Error (MSE) and Root Mean Squared Error (RMSE) for the final XGBoost model, reflecting its predictive accuracy.*
-
-We performed a residual analysis by plotting the residuals against the predicted values to assess the model's fit.
-This allowed us to check whether the XGBoost model accurately captured the patterns in the data and to verify that no major biases or violations of model assumptions were present.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=0.5\textwidth]{../other/graphs/XGBoostResidual.png}
-\caption{XGBoost Model Residuals vs. Predicted Values Plot}
-\label{fig:residual-XGBoost}
-\end{figure}
-
-We also performed cross-validation to ensure the robustness of the XGBoost model.
-This technique allowed us to evaluate the model’s performance across different subsets of the data, helping to confirm that the model generalizes well and is not overly sensitive to any particular portion of the training data.
-
-::: {#tbl-XGBoostCV}
-\begin{center}
-Table 8: XGBoost Cross-Validation Performance
-\end{center}
-
-| Fold  | RMSE       |
-|-------|------------|
-| 1     | 76,693.86  |
-| 2     | 75,692.41  |
-| 3     | 76,193.42  |
-| 4     | 77,365.82  |
-| 5     | 76,552.10  |
-| **Average** | **76,499.52** |
-:::
-
-*Table 8: This table shows the Root Mean Squared Error (RMSE) from cross-validation, demonstrating the model’s performance across different subsets of the data.*
-
-We used matplotlib [@cite_matplotlib] to perform a residual analysis against the predicted values from the XGBoost model.
-We performed cross-validation using cross_val_score function from scikit-learn  [@cite_sklearn] to evaluate the robustness of the model.
+# Models {#sec-models}
 
@@
 
-## Linear Regression Model with Interaction Term
-
-### Linear Regression Results
-
-Table \ref{tbl-coefficients} provides the coefficients for each predictor in the model.
-The coefficients represent the estimated change in the target variable (income) for a one-unit change in the predictor, holding all other variables constant.
-
-The largest positive effect on income is observed for `cat_MORTGAGE`, which has a coefficient of 11,267.90.
-This suggests that individuals with a mortgage tend to have significantly higher incomes compared to those without, when other factors are held constant.
-Additionally, the coefficient for `num__AGE` is 415.91, indicating that income tends to increase with age, though the effect size is relatively small.
-
-On the other hand, some predictors have negative effects on income.
-The variable `cat__SCHLTYPE` (school type) has the largest negative coefficient of -16,320.74, suggesting that attending certain types of schools might be associated with lower income levels.
-Similarly, `cat__GQ` and `cat__MARST` (marital status) have negative coefficients of -7,956.89 and -5,541.39, respectively, implying that certain marital statuses or living arrangements are associated with lower income levels.
-
-Interestingly, the interaction term `num__EDUC_SEX_INTERACTION` has a negative coefficient of -1,832.32, suggesting that the relationship between education and income may differ by gender.
-The negative value could indicate that the combined effect of education and gender has a suppressing effect on income for certain groups.
-
-### Model Performance
-
-As shown in Table \ref{tbl-modelperformance}, the linear regression model explains a small portion of the variance in income, as evidenced by the R-squared value of 0.0775, which indicates that only 7.75% of the variability in income is explained by the predictors in the model.
-The Adjusted R-squared, which adjusts for the number of predictors, is slightly lower at 0.0775, confirming that the inclusion of additional predictors did not drastically improve the model’s explanatory power.
-
-Furthermore, the model’s Mean Squared Error (MSE) is 7,448,800,507, and the Root Mean Squared Error (RMSE) is 86,306.43, suggesting that the model's predictions deviate significantly from the actual income values.
-These performance metrics indicate that the linear regression model has limited predictive accuracy for this dataset.
-
-### Residual Analysis
-
-The residual analysis helps evaluate the linear regression model's performance.
-The residual histogram as shown in Figure \ref{fig:residual-linear1} shows a strong right skew, indicating that while most predictions are close to the actual values, the model struggles with higher incomes.
-The long tail suggests significant under-prediction for some individuals.
-
-In the residuals vs.
-predicted values plot as shown in Figure \ref{fig:residual-linear2}, the residuals exhibit a funnel shape, showing increasing variance as predicted income rises.
-This indicates heteroscedasticity, meaning the model's error increases for higher predicted values.
-Overall, the analysis suggests that the linear model fits well for lower incomes but underperforms for higher-income predictions, implying a need for more complex models to better capture these relationships.
-
-## Random Forest
-
-We employed a Random Forest model to predict income, tuning several hyperparameters to improve model performance.
-The best model was achieved with the following parameters: **max_depth=10**, **min_samples_leaf=4**, **min_samples_split=10**, and **n_estimators=300**.
-As we shown in Table \ref{tbl-randomforestMSE}, the final model had a Mean Squared Error (MSE) of 6,057,426,411.65 and a Root Mean Squared Error (RMSE) of 77,829.47, indicating a substantial improvement in prediction accuracy compared to the linear regression model.
-
-### Feature Importance
-
-The Random Forest model, Table \ref{tbl-randomforestimportance}, also allows for an examination of feature importance, which ranks the variables by their contribution to the model's predictions.
-The most important features in the model are:
-
--   `cat__EDUC_new` (Education level) with an importance score of 0.337, indicating that education is the most significant factor affecting income in this model.
-
--   `cat__IND1990` (Industry) has an importance score of 0.243, suggesting that the industry in which a person works is also a major determinant of income.
-
--   `num__AGE` (Age) comes in third with a score of 0.194, implying that age plays a substantial role in income prediction.
-
-Other variables, such as `cat__SEX` (Sex) and `cat__MARST` (Marital Status), show moderate importance, while factors like `cat__MORTGAGE` and `cat__SCHLTYPE` (School Type) have much smaller impacts.
-
-### Residual Analysis
-
-The residuals from the Random Forest model as shown in Figure \ref{fig:residual-randomforest} were analyzed similarly to the linear model.
-The residuals plot, comparing predicted values with residuals, shows less heteroscedasticity compared to the linear model, suggesting that the Random Forest model handles the data more effectively, particularly for higher income levels.
-However, some degree of variability in the residuals still increases with predicted income, indicating that further tuning or model adjustments may be necessary for optimal performance.
-
-## XGBoost
-
-The XGBoost model, a gradient boosting algorithm, was optimized using a hyperparameter grid search, yielding the best parameters: **learning_rate=0.1**, **max_depth=6**, **min_child_weight=5**, and **n_estimators=300**.
-These parameters were selected to balance model complexity and overfitting control.
-As we shown in Table \ref{tbl-XGBoostMSE}, the final model achieved a Mean Squared Error (MSE) of 5,854,629,848.13 and a Root Mean Squared Error (RMSE) of 76,515.55, making it the best-performing model among the three evaluated.
-
-### Model Performance
-
-The XGBoost model’s Cross-Validation RMSE (Table \ref{tbl-XGBoostCV}) was 76,499.52, which aligns closely with the final RMSE, indicating that the model generalizes well across different data subsets.
-This consistency in performance suggests that XGBoost effectively captures the relationship between predictors and income without overfitting.
-
-### Residual Analysis
-
-The residuals vs.
-predicted values plot as shown in Figure \ref{fig:residual-XGBoost} shows a similar pattern to the Random Forest model, with residuals clustering around zero for lower predicted values but spreading out for higher predicted incomes.
-However, compared to the Random Forest model, the XGBoost residuals show a slightly narrower spread, particularly at the higher income levels, indicating that XGBoost handles high-income predictions somewhat better.
-Nonetheless, some degree of heteroscedasticity remains, as the spread increases with predicted income.
-
 # Discussion {#sec-discussion}
-
-In this analysis, we explored three different models—Linear Regression, Random Forest, and XGBoost—to predict income based on various demographic and categorical factors.
-Each model provided valuable insights into the relationship between income and predictors such as education, age, industry, and gender.
-Among the models, XGBoost demonstrated the best performance in terms of predictive accuracy, as evidenced by its lower RMSE and superior handling of complex data interactions.
-However, this does not mean it is without limitations or that further improvements cannot be made.
-
-## Key Findings
-
-### Feature Importance
-We have found that educational attainment and occupation industry are consistently the most important features in all three models, especially in XGBoost and Random Forest.
-This indicates that higher educational attainment and certain industries are correlated with higher personal income.
-This confirms our original hypothesis that certain demographic factors, such as education, may lead to specialized industry sectors where more income can be generated.
-
-Age is also a useful predictor.
-However, despite the common belief that age is a determining factor when predicting personal income, its effectiveness is not as strong as education or industry.
-This may be caused by the fact that in some industries, your experience may hit a plateau, after which age may not contribute as much to experience as before certain age.
-
-Marital status and sex do affect personal income, but not as significantly as the other features.
-It is great to prove that certain demographical features are not as informative when predicting personal income, suggesting a "fair" distribution of wealth.
-
-### Models
-
-None of the three models used in our study provided an excellent prediction of personal income.
-For the residual analysis, it appears that the models' predictbility deterioate as the self-reported income gets higher.
-REsidual variance increased as the predicted income increased in the Random Forest and XGBoost model, suggesting potential heteroscedasticity.
-Our educated guess is that extra-high income may come from various sources, and simplifying them into a simple model like this is not a reasonable solution.
-
-Among the three models, XGBoost has the highest accuracy overall.
-We believe that it can capture more complex relationships and interactions among the variables than the other two models.
-However, the high MSE for extra-high income individuals indicates that the model still needs to be refined more to achieve an accurate prediction.
-
-Linear Regression has underperformed relative to the other two models.
-We speculate that simplicity in modeling linear relationships did not provide sufficient information to capture the non-linear patterns underlying our data, especially where multiple features interact to influence income, as suggested in @sec-afr.
-
-## Weaknesses and Limitations
-
-One major limitation encountered throughout the analysis was the skewness of the income distribution.
-The income data is heavily left-skewed, with most individuals earning relatively low amounts and a few individuals earning substantially higher incomes.
-This skewness likely contributed to the residual patterns we observed, particularly in the linear and Random Forest models.
-
-We attempted to address this by applying a logarithmic transformation to income in an effort to normalize the data.
-However, this transformation did not lead to substantial improvements in model performance.
-Even after the transformation, the residual patterns remained, indicating that the underlying relationships between predictors and income might require more sophisticated modeling approaches than simple transformations can provide.
-
-Another potential solution—though not ideal—would be to exclude extreme income values from the analysis.
-Removing high-income outliers might reduce the heteroscedasticity issue and result in a better fit for the majority of the population.
-However, excluding high-income individuals would also mean losing valuable information about the upper end of the income distribution, which is crucial for a complete analysis.
-Before taking this step, it would be necessary to justify it based on domain-specific knowledge or further data exploration.
-
-## Next Steps
-
-Looking forward, there are several potential avenues for further research and model improvement:
-
--   While we focused on standard predictors like age, education, and industry, additional variables or interaction terms might improve model performance.
-For example, interactions between age and education, or industry and gender, could capture more nuanced relationships.
-
--    Instead of removing high-income individuals, more sophisticated outlier treatment techniques could be applied.
-For instance, applying quantile regression or using robust regression models might help better model extreme values without distorting the broader data patterns.
-
--   The skewness in income distribution poses a significant challenge for modeling.
-Exploring techniques like synthetic data generation for high-income individuals or applying weighting schemes to account for data imbalance might help the models better capture the variance across all income levels.
-
-In conclusion, while XGBoost provided the most accurate results, there is room for improvement in handling skewed data and better modeling high-income individuals.
-Future work will focus on addressing these issues while exploring more advanced modeling techniques and feature engineering to enhance predictive performance further.
-
-\newpage
-
-# Appendix
-
-## Data Cleaning {#sec-datacleaning}
-
-In the data cleaning process, we first examined all required variables and removed any observations with missing values (N/A).
-Notably, the variable `CITY`, which we initially intended to include in the analysis, was excluded due to its high proportion of missing values, which compromised data completeness.
-
-For the response variable, personal total income (INCTOT), we identified several non-informative coded values, such as `0000000 = None`, `9999998 = Unknown`, and `-009995 = -$9,900 (1980)` and so on.
-These values lack meaningful information for analysis, so we removed any records containing such entries to maintain the integrity of the income data.
-
-In terms of predictor variables, we found that several included codes like "not applicable" or "does not exist," which represent cases without relevant data for our analysis.
-These values were also removed to ensure only relevant data points were used in modeling.
-
-Furthermore, we simplified the `EDUC` variable (education level), which originally included highly granular categories (e.g., fifth grade, sixth grade, second year of college, fifth year of college).
-Such detailed classifications were unnecessary for our purposes, so we re-coded `EDUC` into a new variable, `EDUC_new`, with broader categories:
-
--   'No Schooling': 0,
--   'Nursery School': 1,
--   'Primary School': 2,
--   'Middle School': 3,
--   'College 1-4 Years': 4,
--   'College 5+ Years': 5
-
-This simplified categorization enhances interpretability and usability in the analysis.
-By conducting these data cleaning steps, we produced a refined and consistent dataset, ready for accurate and reliable modeling.
 

### 2024-12-06 paper/paper.qmd
More graphs saved from the models' results. Model, Results completed.

@@
 #| message: false
-#| fig-pos: "H"
 
@@
 
+## Recurrent Neural Network (RNN) {#sec-rnn}
+First published in [@elman1990finding] in 1990, RNN is a type of neural network that is designed to handle sequential data.
+Different from a feedforward neural network, RNN feeds the output from one step back as input for the next step, enabling RNN to "remember" information from the previous inputs, as described in [@GeeksforGeeks_2024].
+This key mechanism makes RNN well-suited for time-series predictions and other sequential tasks.
+In this project, an RNN model was trained to predict Bitcoin's price movement direction for the 1-minute, 1-hour, and 1-day datasets.
+The features included lagged price differences (open_diff, high_diff, low_diff, close_diff), volume, a missing-data indicator (was_missing) for the 1-minute and 1-hour datasets, and the previous price movement direction (direction_t-1).
+The target variable was the future price movement direction (direction_t+1), which was coded into three classes: -1 (decrease), 0 (no change), and 1 (increase), as mentioned in @sec-featureengineering.
+In this project, the RNN has two simple RNN layers, with 128 and 64 nodes, respectively; a fully connected layer with 64 nodes; and the output layer uses softmax as its activation function, with three possible predictions: -1 (decrease), 0 (no change), and 1 (increase).
+Dropout was performed at all layers before the output layer to prevent overfitting.
+The loss function was cross-entropy.
+
+## Long Short-Term Memory (LSTM) {#sec-lstm}
+Originally published in [@hochreiter1997long] in 1997, LSTM is a special type of RNN, and it is designed to handle both short- and long-term dependencies in time series data, as stated in [@ghojogh2023recurrent].
+Leaving out all the technical details, the primary function of LSTM is to retain important historical information, thus making it an excellent model for predicting sequential data, like financial forecasting and NLP.
+In this project, an LSTM model was trained to predict Bitcoin's price movement direction at different time intervals.
+Using the LSTM trained on the 1-minute dataset as an example, the features used for prediction are the lagged differences of price variables (open_diff, high_diff, low_diff, close_diff), volume, the missing-data indicator (was_missing), and the direction of the price movement at the previous timestamp (direction_t-1).
+These features were normalized using a MinMaxScaler offered in [@pedregosa2011scikit] to improve the model's performance.
+The dataset was transformed into sequences with a time step of 10 (10 observations in 10 minutes) to allow LSTM to learn short-term patterns over short historical periods.
+As for the model itself, the LSTM model has one LSTM layer with 128 nodes with a ReLU activation function, two fully connected layers with 64 nodes per layer with random dropout to prevent overfitting, and a binary output layer with a sigmoid activation function.
+The loss function was cross-entropy.
+
+## Extreme Gradient Boosting (XGB) {#sec-xgb}
+According to [@chen2016xgboost], XGB is a highly optimized and efficient implementation of the gradient boosting framework.
+In [@shwartz2022tabular], the authors stated that "XGBoost is widely used for structured and tabular data due to its flexibility, performance, and ability to handle missing data effectively." Despite not being explicitly developed for sequential data, XGB is a commonly used machine learning model for both regression and classification with excellent performance, according to [@Mukherjee_2023].
+In this project, XGB classifiers were used to predict Bitcoin's price movement direction for all three datasets.
+Unlike the previous two sequential models (LSTM and RNN), XGB works on tabular data without requiring sequential dependencies.
+The features included lagged price differences (open_diff, high_diff, low_diff, close_diff), volume, a missing-data indicator (was_missing) for the 1-minute and 1-hour datasets, and the previous price movement direction (direction_t-1).
+The target variable was the future price movement direction (direction_t+1), which was coded into three classes: -1 (decrease), 0 (no change), and 1 (increase), as mentioned in @sec-featureengineering.
+The XGB model itself has 100 estimators to balance training speed and model complexity, a maximum depth of 10, and a 0.1 learning rate.
+
 # Results {#sec-results}
+
+## Performance 
+This subsection compares the performance of the three models across each dataset.
+Key metrics, including accuracy, F1-score, and confusion matrices, are used to evaluate and contrast their effectiveness.
+
+### RNN
+RNN performs the worst when it is applied to predict a 1-minute interval dataset.
+The overall accuracy is only 0.41, and the F-1 scores are 0.51, 0.20, and 0.35, respectively, for the prediction of a decrease, no change, and an increase.
+When used on the 1-hour dataset, RNN performs significantly better, with an overall 0.52 accuracy and 0.52 F-1 scores for both increase and decrease.
+Lastly, RNN also performs well on the 1-day dataset, with an overall 0.52 accuracy and 0.32 and 0.63 F-1 scores for predicting a decrease and increase.
+The confusion matrixes of RNN on these three different time intervals are shown in @fig-figure9, @fig-figure10, and @fig-figure11.
+
+```{python}
+
+#| echo: false
+#| label: fig-figure9
+#| fig-cap: "Confusion matrix for RNN on 1-minute dataset"
+#| warning: false
+#| message: false
+
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
+
+img = mpimg.imread('../other/graphs/rnn_confusion_1m.png')
+plt.figure(figsize=(10, 6))
+plt.imshow(img)
+plt.axis('off')
+plt.show()
+
+```
+
+```{python}
+
+#| echo: false
+#| label: fig-figure10
+#| fig-cap: "Confusion matrix for RNN on 1-hour dataset"
+#| warning: false
+#| message: false
+
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
+
+img = mpimg.imread('../other/graphs/rnn_confusion_1h.png')
+plt.figure(figsize=(10, 6))
+plt.imshow(img)
+plt.axis('off')
+plt.show()
+
+```
+
+```{python}
+
+#| echo: false
+#| label: fig-figure11
+#| fig-cap: "Confusion matrix for RNN on 1-day dataset"
+#| warning: false
+#| message: false
+
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
+
+img = mpimg.imread('../other/graphs/rnn_confusion_1d.png')
+plt.figure(figsize=(10, 6))
+plt.imshow(img)
+plt.axis('off')
+plt.show()
+
+```
+
+### LSTM
+LSTM did not work as intended.
+All of them only predict a decrease for the next timestamp.
+For example, for the 1-minute dataset, LSTM produced an overall accuracy of 0.48, and its confusion matrix is shown in @fig-figure12.
+The other statistics will not be provided here in detail because they are useless, considering the model is not making any meaningful prediction.
+For the details of LSTM models, please read about them in the actual model script, called "06-model_data.ipynb", located under the "scripts" folder.
+
+```{python}
+
+#| echo: false
+#| label: fig-figure12
+#| fig-cap: "Confusion matrix for LSTM on 1-hour dataset"
+#| warning: false
+#| message: false
+
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
+
+img = mpimg.imread('../other/graphs/lstm_confusion_1h.png')
+plt.figure(figsize=(10, 6))
+plt.imshow(img)
+plt.axis('off')
+plt.show()
+
+```
+
+### XGB
+Opposite to RNN, XGB performs well when it is applied to predict a 1-minute interval dataset.
+The overall accuracy is only 0.53, and the F-1 scores are 0.58 and 0.46 for the prediction of a decrease and an increase.
+When used on the 1-hour dataset, XGB produced an overall 0.53 accuracy and 0.51 and 0.55 for the prediction of a decrease and an increase.
+Lastly, XGB performs slightly worse on the 1-day dataset, with an overall 0.50 accuracy and 0.49 and 0.52 F-1 scores for predicting a decrease and increase.
+The confusion matrixes of RNN on these three different time intervals are shown in @fig-figure13, @fig-figure14, and @fig-figure15.
+
+```{python}
+
+#| echo: false
+#| label: fig-figure13
+#| fig-cap: "Confusion matrix for XGB on 1-minute dataset"
+#| warning: false
+#| message: false
+
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
+
+img = mpimg.imread('../other/graphs/XGB_confusion_1m.png')
+plt.figure(figsize=(10, 6))
+plt.imshow(img)
+plt.axis('off')
+plt.show()
+
+```
+```{python}
+
+#| echo: false
+#| label: fig-figure14
+#| fig-cap: "Confusion matrix for XGB on 1-hour dataset"
+#| warning: false
+#| message: false
+
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
+
+img = mpimg.imread('../other/graphs/XGB_confusion_1h.png')
+plt.figure(figsize=(10, 6))
+plt.imshow(img)
+plt.axis('off')
+plt.show()
+
+```
+
+```{python}
+
+#| echo: false
+#| label: fig-figure15
+#| fig-cap: "Confusion matrix for XGB on 1-day dataset"
+#| warning: false
+#| message: false
+
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
+
+img = mpimg.imread('../other/graphs/XGB_confusion_1d.png')
+plt.figure(figsize=(10, 6))
+plt.imshow(img)
+plt.axis('off')
+plt.show()
+
+```
+
+### Model Performance Comparison
+Among the three tested models, XGB produced the best overall accuracy, especially when predicting on the 1-minute and 1-hour interval datasets, showing its strength in handling the data effectively.
+In contrast, LSTM performed the worst, failing to make meaningful predictions and only predicting a decrease for all instances.
+This limited its performance in this study.
+While RNN struggled with the 1-minute dataset, it performed better on the 1-hour and 1-day datasets and even achieved the highest F1-score of 0.63 for predicting an increase on the 1-day dataset.
+This suggests that RNN may be better suited for datasets with larger time intervals, where sequential dependencies are more influential.
+Overall, the results indicate that XGB offers a more reliable balance of performance and efficiency.
+In contrast, the sequential models (RNN and LSTM) require further optimization or adjustments to deliver competitive results.
+
+## Training Effort
+XGB trains the fastest, followed by RNN, with LSTM taking the longest to train when using only a CPU.
+This project used a 16-core Intel Core i7-10700KF 3.80GHz CPU to train all the models.
+Using the 1-minute dataset as an example, XGB took only 8 minutes to complete training, RNN took 2 and 40 minutes, while LSTM used 10 and 31 minutes.
+These results confirmed the efficiency of XGB, which is particularly well-suited for individuals with limited computational resources.
+RNN, though slower, demonstrates a reasonable trade-off between training time and predictive performance, particularly for datasets with longer intervals.
+On the other hand, LSTM has the most prolonged training duration and least optimal performance.
+This suggests that when training LSTM models, better optimization or alternative approaches can be beneficial without access to GPUs or distributed computing resources.
+This extreme difference in training times demonstrates the practical implications of model selection for computationally constrained traders.
 

### 2024-12-06 paper/paper.qmd
Discussion completed. LLM usage updated

@@
 # Discussion {#sec-discussion}
+In this analysis, we explored three models—RNN, LSTM, and XGBoost—to predict the direction of Bitcoin price movement with freely available information.
+Among the models, XGBoost demonstrated the best overall performance in terms of predictive accuracy and the shortest training time.
+RNN also predicts well when applied to longer time intervals (1 day).
+More details are discussed in the following subsections.
+
+## Key Findings
+
+### The Strength of XGB
+XGB has demonstrated the best overall accuracy across the 1-minute and 1-hour datasets.
+This confirms XGB's superior ability to handle tabular data with limited resources, as mentioned in [@bernson2024xgboost].
+Moreover, XGB required the shortest training time, making it time-efficient for model training and retraining.
+
+### RNN's Performance on Longer Time Duration
+Although RNN performed poorly on the 1-minute dataset, it excelled at prediction Bitcoin's price movement at 1-hour and 1-day time interval.
+It achieved the highest F1-score in this study for predicting price increases in the 1-day dataset.
+This indicates that RNNs might be better suited for datasets where sequential patterns are more pronounced over longer durations.
+
+### The Limitations of LSTM
+LSTM failed to make any meaningful predictions, and it only gave price decrease predictions in this study.
+Additionally, LSTM required the longest training time, with minimal return in predictive performance, making it unusable without further optimization and tuning.
+
+### Training Efficiency Comparison
+XGB significantly outperformed both RNN and LSTM in their training speeds, completing training in mere minutes compared to hours for the other models.
+This shows XGB's computational efficiency, which can be useful for traders or researchers with limited computational power.
+
+## Limitations
+While the study provided valuable insights, several limitations are also recognized and must be addressed.
+Firstly, due to time constraints, the models were not fully optimized.
+This may have led to the suboptimal performance of RNN (on the 1-minute dataset) and LSTM models.
+Secondly, including "0" as an outcome class for RNN and LSTM models may have diluted the predictive power without providing additional helpful information to assist trading.
+Removing this outcome and focusing on price increases or decreases might provide more precise insights for investors.
+Thirdly, all training was conducted using a CPU.
+This limitation may impact the training time for RNN and LSTM because GPU training has much higher parallelization capacities, which can be optimized when training deep learning models, as mentioned in [@buber2018performance].
+Fourthly, only basic features were used in the models.
+Although the goal of this project is to use readily available features to predict Bitcoin's price movement direction, there are other features that are also easy to acquire, such as its hash rate, which may be very effective when predicting the direction of cryptocurrency's price, according to [@king2024blockchain].
+Lastly, the dataset was sourced from Binance and Binance only, and this may limit the data's quality and the models' generalizability.
+Cross-verifying the data with other sources might provide more robust results.
+
+## Next Steps
+To address the above limitations and further improve the models, the next steps of this research are to tune the hyperparameters of the models and improve their performance, especially for RNN and LSTM; refine the outcome classes and remove "0" (no change) from the output list to focus on providing actionable outcomes; add more features and train the deep learning models on GPU; and lastly, combine data from multiple sources to make the models and results more robust.
 
